---
title: 'Oocyte Alternative Splicing Analysis'
Subtitle: "FMN2KO & SPIRE1/2KO"
author: "Andrés Gordo Ortiz @Al Jord Lab CRG"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    code_folding: hide
    code_download: false
    df_print: paged
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true

editor_options:
  markdown:
    wrap: 72


header-includes:
  - '<link href="https://fonts.googleapis.com/css2?family=Roboto:wght@400;700&display=swap" rel="stylesheet">'
  - '<style>
        #logos {
            display: flex;
            justify-content: center;
            align-items: center;
            margin-bottom: 20px;
        }
        #logos img {
            height: 80px;
            margin: 0 15px;
        }
     </style>'
---



```{css, echo=FALSE}
/* Minimal and clean custom CSS */
body {
  font-family: 'Roboto', sans-serif;
  color: #333; /* Neutral dark gray for body text */
  background-color: #f9f9f9; /* Light background */
  margin: 0;
  padding: 0;
  width: 100%; /* Full width */
}

.container {
  max-width: 100%; /* Use full width of the screen */
  padding: 20px; /* Optional: add padding for better readability */
}

h1, h2, h3, h4, h5, h6 {
  font-family: 'Roboto', sans-serif;
  font-weight: bold;
}

/* Header Colors */
h1 {
  font-size: 2.5em;
  color: #5E14C4; /* Main headers in purple */
}

h2 {
  font-size: 2em;
  color: #4A10A1; /* Darker purple for secondary headers */
}

h3 {
  font-size: 1.75em;
  color: #3B0D7E; /* Even darker purple for tertiary headers */
}

p {
  text-align: justify;
  line-height: 1.6;
  margin: 0 0 1em;
}

/* Styling for subtitle, authors, and date */
.subtitle, .author, .date {
  color: #5E14C4;
  font-size: 1.25em;
  text-align: center;
  margin-bottom: 0.5em;
}

/* Link styling */
a {
  color: #5E14C4;
  text-decoration: none;
}

a:hover {
  text-decoration: underline;
}

/* Table styling */
table {
  border-collapse: collapse;
  width: 100%;
  margin: 20px 0;
}

table th, table td {
  border: 1px solid #ddd;
  padding: 10px;
}

table th {
  background-color: #5E14C4;
  color: white;
}

table tr:nth-child(even) {
  background-color: #f9f9f9;
}

table tr:hover {
  background-color: #f1f1f1;
}


```



```{r setup, include=FALSE}
# Ensure a clean environment and load required libraries.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)


# Load required libraries
library("betAS")
library("ggplot2")
library("plotly")
library("dplyr")
library("tidyverse")
library("cowplot")
library("DT")
library("paletteer")
library("ggupset")
library("showtext")
library("ggtext")
library("readxl")
library("biomaRt")
library("patchwork")
library("org.Mm.eg.db")
library("org.Hs.eg.db")
library("enrichplot")
library("DOSE")
library("clusterProfiler")
library("ComplexHeatmap")
library("enrichR")
library("colorRamp2")
library("dendextend")

# Register Google fonts
showtext_auto()
font_add_google("Roboto", "roboto")
font="roboto"

```

# Introduction to this Analysis

## Part 1: Upstream Analysis

The upstream analysis is performed on the CRG cluster using the Bash command line. This phase involves:

1. **Data Acquisition**:
  - Downloading `*.fastq.gz` files of these studies:
    - **FMN2 DKO Oocytes**
    - **SPIRE1/2 DKO Oocytes**
  - Merging technical replicates for superior sequencing depth. RNA Splicing analysis ussually requires a minimum of 50 million reads per biological sample.

2. **Alignment**:
   - Aligning the sequences to the latest mouse genome release (mm10) using **Bowtie2** through the [Vast-Tools align function](https://link.springer.com/protocol/10.1007/978-1-0716-2521-7_7).

3. **Event Detection**:
   - Importing inclusion tables generated by Vast-Tools into RStudio. Vast-Tools does **not** find new splicing events, relaying on a manually curated database, [VASTDB](https://vastdb.crg.eu/), with more than 600.000 splicing events. The end result is an `Inclusion_table.tab`, which contains the splicing events as rows and the samples as columns. Splicing is quantified using the Percentage Spliced-In (PSI) parameter, which shows the percentage at which a given splicing event occurs across all sequencing reads of the gene. The database covers the following splicing events:
   
     - **Exons (EX)**: Cassette sequences that are either included or excluded in the mature mRNA.
     - **Introns (IN)**: Introns that are either retained or spliced out in the mature mRNA.
     - **Alternative 5' Splice Sites (Alt5)**: Exons that are spliced at different 5' sites (donor sites).
     - **Alternative 3' Splice Sites (Alt3)**: Exons that are spliced at different 3' sites (acceptor sites).
     - **Microexons (MIC)**: Exons that are shorter than 27 nucleotides.

---

## Part 2: Statistical Analysis

State-of-Art splicing analysis tools such as the [betAS package](https://rnajournal.cshlp.org/content/30/4/337) employ a combination of statistical testing and simulations on the *beta* distribution to calculate the delta PSI (the difference in splicing between two conditions) and the False Discovery rate (FDR, or adjusted p-value) respectively. This is computationally demanding for such a large dataset. For that reason we focused first on analysing *alternative* splicing events (that is, events with a PSI different that *0* or *100*--not always active, not always inactive) which constitute ~25% of all events in both *FMN2* and *Spire* studies. With that first analysis done (which can be accessed completely in the file `alternative_splicing_fmn2_spire.html`) we fine-tuned the code and used the complete dataset. As expected, almost all significant events found in the smaller subset are found using it fully (*see Table x*). The specific steps taken for these analysis were.


1. **Statistical Testing**:
   - Use the [betAS package](https://rnajournal.cshlp.org/content/30/4/337) in R to perform Beta Distribution simulations to obtain:
     - **False Discovery Rate (FDR)**
     - **delta Percentage Spliced-in (dPSI)**

2. **Filtering Criteria**: Significant events are defined by
     - FDR <= 0.01
     - |dPSI| >= 0.1 (out of 1. That is, at least a 10% increase or decrease in PSI)


---


```{r inclusion tables}

fmndko_data <- getDataset(pathTables = paste0(getwd(),"/inclusion_tables/fmndko_INCLUSION_LEVELS_FULL-mm10.tab"), tool = "vast-tools")
fmndko_events <- getEvents(fmndko_data, tool = "vast-tools") # Extract alternative splicing events

fmndko_exons <- filterEvents(fmndko_events, types = c("C1", "C2", "C3", "S", "MIC"), N = 2)
fmndko_introns <- filterEvents(fmndko_events, types = c("IR"), N = 2)
fmndko_alt <- filterEvents(fmndko_events, types = c("Alt5", "Alt3"), N = 2)



spire_data <- getDataset(pathTables = paste0(getwd(),"/inclusion_tables/spiredko_INCLUSION_LEVELS_FULL-mm10.tab"), tool = "vast-tools")
spire_events <- getEvents(spire_data, tool = "vast-tools") # Extract alternative splicing events

spire_exons <- filterEvents(spire_events, types = c("C1", "C2", "C3", "S", "MIC"), N = 2)
spire_introns <- filterEvents(spire_events, types = c("IR"), N = 2)
spire_alt <- filterEvents(spire_events, types = c("Alt5", "Alt3"), N = 2)


```


## Metadata { .tabset}

### FMN2DKO

```{r metadata fmn2dko}
# Load metadata file containing sample information
metadata_fmn2dko<-data.frame(samples=c("SRR6026682_SRR6026683_SRR6026684_merged","SRR6026685_SRR6026686_SRR6026687_merged","SRR6026688_SRR6026689_SRR6026690_merged","SRR6026691_SRR6026692_SRR6026693_merged"),condition=c("control","control","fmndko","fmndko"))
DT::datatable(metadata_fmn2dko, options = list(pageLength = nrow(metadata_fmn2dko), scrollX = TRUE))
```

### Spire

```{r metadata tao}
# Load metadata file containing sample information
metadata_spire<-data.frame(samples=c("Oocytes_FG_Spire12_Cont_a","Oocytes_FG_Spire12_Cont_b","Oocytes_FG_Spire12_Cont_c","Oocytes_FG_Spire12_DKO_a","Oocytes_FG_Spire12_DKO_b","Oocytes_FG_Spire12_DKO_c"), condition=rep(c("control", "spiredko"), each=3))
DT::datatable(metadata_spire, options = list(pageLength = nrow(metadata_spire), scrollX = TRUE))
```


---



# Total Splicing Events

This plot shows the total number of splicing events found during the **mapping**, alongside the proportions of each type of event. It is important to know that *vast-tools* does not find new events, but it identifies those manually curated in the database. The total number is the same for both datasets, as it is the total number of events registered and curated in ***VASTDB***, hence the *Complete Version.* 


```{r splicing events barplot stacked}

# Adjusted version of the script
splicing_events <- tibble(
  event_type = names(spire_events$EventsPerType),
  spire = spire_events$EventsPerType,
  fmndko = fmndko_events$EventsPerType
)

splicing_events_long <- splicing_events %>%
  pivot_longer(cols = c(spire, fmndko),
               names_to = "study", values_to = "count") %>%
  mutate(event_type = ifelse(event_type %in% c("C1", "C2", "C3", "ANN", "S", "MIC"), "EX", event_type)) %>%
  group_by(study, event_type) %>%
  summarize(count = sum(count, na.rm = TRUE), .groups = "drop")

# Proportion bar plot
plot_proportion <- ggplot(splicing_events_long, aes(fill = event_type, y = count, x = study)) +
  geom_bar(position = "fill", stat = "identity") +
  labs(
    title = "<b style='font-size:18px;'>Proportion of Splicing Events by Study</b>",
    x = NULL,
    y = "Proportion of Events",
    fill = "Event Type"
  ) +
  theme_minimal(base_family = font) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_markdown(),
    panel.background = element_rect(fill = "gray98"),
    panel.grid = element_line(color = "gray90", size = 0.5)
  ) +
  scale_fill_paletteer_d("MetBrewer::Hokusai3")

splicing_events_long_sum<-splicing_events_long %>%
  dplyr::group_by(study) %>%
  summarize(count = sum(count))


# Absolute counts plot
plot_absolute <- ggplot(splicing_events_long_sum, aes(y = count, x = study)) +
  geom_point(color = "#C70039", size = 3) +
  geom_text(aes(label = count), vjust = -1, size = 4, check_overlap = TRUE) +
  labs(
    x = "Study",
    y = "Number of Events",
    caption = paste0("Created by AG on ", Sys.Date()), size=3
  ) +
  theme_minimal(base_family = font) +
  theme(
    legend.position = "none",
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.text.x = element_text( size=15),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray80", size = 0.5),
    panel.grid.minor = element_line(color = "gray95", size = 0.3)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.2)))


# Combine plots
combined_plot <- plot_proportion / plot_absolute +
  plot_layout(heights = c(1, 0.5)) &
  theme(plot.margin = margin(5,5, 0, 5))

# Display combined plot
combined_plot
```

```{r general stats}
vastdb_events<-read.table("EVENT_INFO-mm10.tab", header = TRUE, sep = "\t")
bin_exons <- function(n) {
  if (is.na(n)) return(NA) # Handle missing values
  else if (n < 6) return("1-5")
  else if (n < 9) return("6-10")
  else if (n < 21) return("11-20")
  else return("21+")
}

#subset vastdb events if they contin the word "EX" in the column EVENT
vastdb_exons<-vastdb_events[grep("EX", vastdb_events$EVENT),]
# Add exon group information
apply_bin_exons <- function(item) {
  gene_counts <- table(item$GENE) # Count exons per gene
  item$exon_group <- sapply(item$GENE, function(gene) bin_exons(gene_counts[gene]))
  item
}

vastdb_exons<-apply_bin_exons(vastdb_exons)

#subset vastdb events if they contin the word "EX" in the column EVENT
vastdb_introns<-vastdb_events[grep("IN", vastdb_events$EVENT),]
# Add exon group information
apply_bin_exons <- function(item) {
  gene_counts <- table(item$GENE) # Count exons per gene
  item$intron_group <- sapply(item$GENE, function(gene) bin_exons(gene_counts[gene]))
  item
}

vastdb_introns<-apply_bin_exons(vastdb_introns)

```

# Inclusion or Exclusion Plots { .tabset}

These plots show the distribution of the splicing change (dPSI) comparing the KO mutant vs the control. If there is a tendency towards inclusion or exclusion, the bell-shaped curve will be shifted to the right or left, respectively. First, I calculated the mean of each event per condition and then I calculated the difference between means for each event. The subtitle shows the Wilcoxon test assessing this possible assymetry. The area under the curve for both sides (>0 and <0) is shown numerically. Only events with a dPSI greater or lower than 0 are shown for clarity.

## Exons 

### Spire

```{r inclusionumber of exons spire}
# ---- Prepare Data -----------------------------------------------------------
# Remove any NA rows (adjust as needed)
spire_exons_df <- na.omit(spire_exons$PSI)

# Function to compute the PSI difference for each row (KO - Control)
compute_difference <- function(df) {
  avg_KO <- rowMeans(df[, 10:12])
  avg_control <- rowMeans(df[, 7:9])
  avg_KO - avg_control
}

# Calculate the difference and remove any additional NAs
spire_exons_df$difference <- compute_difference(spire_exons_df)
spire_exons_df <- na.omit(spire_exons_df)

# Create a group variable based on whether the difference is below or above zero
spire_exons_df$group <- ifelse(spire_exons_df$difference < 0, "Skipped", "Included")

# Perform a one-sample Wilcoxon signed-rank test (null hypothesis: median = 0)
symmetry_test <- wilcox.test(spire_exons_df$difference, mu = 0)

# Compute the percentages (areas under the density curve)
pct_below <- round(mean(spire_exons_df$difference < 0) * 100, 1)
pct_above <- round(mean(spire_exons_df$difference > 0) * 100, 1)

# ---- Plot -------------------------------------------------------------------

plot_psi_distribution <- ggplot(
  spire_exons_df[abs(spire_exons_df$difference) > 0,], 
  aes(y = ..density.., x = difference, fill = group)
) +
  geom_histogram(bins = 200, position = "identity", alpha = 0.9) +
  labs(
    title = "PSI Difference Distribution by Exon Group",
    subtitle = paste("One-sample Wilcoxon test p-value for Symmetry:", format(symmetry_test$p.value, digits = 3)),
    x = "ΔPSI (KO - Control)",
    y = "Density",
    fill = "Splicing Direction",
    caption = paste0("Created by AG on ", Sys.Date())  # Display quantiles in caption
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    legend.position = "right",
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.text.x = element_text(size = 15),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray80", size = 0.5),
    panel.grid.minor = element_line(color = "gray95", size = 0.3),
    strip.text = element_text(size = 14, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) +
  # Set the x-axis breaks
  scale_x_continuous(
    breaks = c(-100, -75, -50, -25, 0, 25, 50, 75, 100)
  ) +
  # Annotate the percentages on the left and right sides of the plot
  annotate("text", x=-65, y=0.2, hjust=0, vjust=0, label = paste0("dPSI<0: ",pct_below, "%"), color = "blue", size = 7) +
  annotate("text", x=65, y=0.2, hjust=1, vjust=0,
           label = paste0("dPSI>0: ",pct_above, "%"), color = "red", size = 7) +
  scale_fill_brewer(palette = "Set1")

print(plot_psi_distribution)

```



### FMNDKO

```{r inclusionumber of exons fmndko}
# ---- Prepare Data -----------------------------------------------------------
# Remove any NA rows (adjust as needed)
fmndko_exons_df <- na.omit(fmndko_exons$PSI)

# Function to compute the PSI difference for each row (KO - Control)
compute_difference <- function(df) {
  avg_KO <- rowMeans(df[, 9:10])
  avg_control <- rowMeans(df[, 7:8])
  avg_KO - avg_control
}

# Calculate the difference and remove any additional NAs
fmndko_exons_df$difference <- compute_difference(fmndko_exons_df)
fmndko_exons_df <- na.omit(fmndko_exons_df)

# Create a group variable based on whether the difference is below or above zero
fmndko_exons_df$group <- ifelse(fmndko_exons_df$difference < 0, "Skipped", "Included")

# Perform a one-sample Wilcoxon signed-rank test (null hypothesis: median = 0)
symmetry_test <- wilcox.test(fmndko_exons_df$difference, mu = 0)

# Compute the percentages (areas under the density curve)
pct_below <- round(mean(fmndko_exons_df$difference < 0) * 100, 1)
pct_above <- round(mean(fmndko_exons_df$difference > 0) * 100, 1)

# ---- Plot -------------------------------------------------------------------

plot_psi_distribution <- ggplot(
  fmndko_exons_df[abs(fmndko_exons_df$difference) > 0,], 
  aes(y = ..density.., x = difference, fill = group)
) +
  geom_histogram(bins = 200, position = "identity", alpha = 0.9) +
  labs(
    title = "PSI Difference Distribution by Exon Group",
    subtitle = paste("One-sample Wilcoxon test p-value for Symmetry:", format(symmetry_test$p.value, digits = 3)),
    x = "ΔPSI (KO - Control)",
    y = "Density",
    fill = "Splicing Direction",
    caption = paste0("Created by AG on ", Sys.Date())  # Display quantiles in caption
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    legend.position = "right",
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.text.x = element_text(size = 15),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray80", size = 0.5),
    panel.grid.minor = element_line(color = "gray95", size = 0.3),
    strip.text = element_text(size = 14, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) +
  # Set the x-axis breaks
  scale_x_continuous(
    breaks = c(-100, -75, -50, -25, 0, 25, 50, 75, 100)
  ) +
  # Annotate the percentages on the left and right sides of the plot
  annotate("text", x=-65, y=0.1, hjust=0, vjust=0, label = paste0("dPSI<0: ",pct_below, "%"), color = "blue", size = 7) +
  annotate("text", x=65, y=0.1, hjust=1, vjust=0,
           label = paste0("dPSI>0: ",pct_above, "%"), color = "red", size = 7) +
  scale_fill_brewer(palette = "Set1")

print(plot_psi_distribution)

```

## Introns

### Spire

```{r inclusionumber of introns spire}
# ---- Prepare Data -----------------------------------------------------------
# Remove any NA rows (adjust as needed)
spire_introns_df <- na.omit(spire_introns$PSI)

# Function to compute the PSI difference for each row (KO - Control)
compute_difference <- function(df) {
  avg_KO <- rowMeans(df[, 10:12])
  avg_control <- rowMeans(df[, 7:9])
  avg_KO - avg_control
}

# Calculate the difference and remove any additional NAs
spire_introns_df$difference <- compute_difference(spire_introns_df)
spire_introns_df <- na.omit(spire_introns_df)

# Create a group variable based on whether the difference is below or above zero
spire_introns_df$group <- ifelse(spire_introns_df$difference < 0, "Skipped", "Included")

# Perform a one-sample Wilcoxon signed-rank test (null hypothesis: median = 0)
symmetry_test <- wilcox.test(spire_introns_df$difference, mu = 0)

# Compute the percentages (areas under the density curve)
pct_below <- round(mean(spire_introns_df$difference < 0) * 100, 1)
pct_above <- round(mean(spire_introns_df$difference > 0) * 100, 1)

# ---- Plot -------------------------------------------------------------------

plot_psi_distribution <- ggplot(
  spire_introns_df[abs(spire_introns_df$difference) > 0,], 
  aes(y = ..density.., x = difference, fill = group)
) +
  geom_histogram(bins = 200, position = "identity", alpha = 0.9) +
  labs(
    title = "PSI Difference Distribution by Exon Group",
    subtitle = paste("One-sample Wilcoxon test p-value for Symmetry:", format(symmetry_test$p.value, digits = 3)),
    x = "ΔPSI (KO - Control)",
    y = "Density",
    fill = "Splicing Direction",
    caption = paste0("Created by AG on ", Sys.Date())  # Display quantiles in caption
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    legend.position = "right",
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.text.x = element_text(size = 15),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray80", size = 0.5),
    panel.grid.minor = element_line(color = "gray95", size = 0.3),
    strip.text = element_text(size = 14, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) +
  # Set the x-axis breaks
  scale_x_continuous(
    breaks = c(-100, -75, -50, -25, 0, 25, 50, 75, 100)
  ) +
  # Annotate the percentages on the left and right sides of the plot
  annotate("text", x=-65, y=0.2, hjust=0, vjust=0, label = paste0("dPSI<0: ",pct_below, "%"), color = "blue", size = 7) +
  annotate("text", x=65, y=0.2, hjust=1, vjust=0,
           label = paste0("dPSI>0: ",pct_above, "%"), color = "red", size = 7) +
  scale_fill_brewer(palette = "Set1")

print(plot_psi_distribution)

```

### FMNDKO

```{r inclusionumber of introns fmndko}
# ---- Prepare Data -----------------------------------------------------------
# Remove any NA rows (adjust as needed)
fmndko_introns_df <- na.omit(fmndko_introns$PSI)

# Function to compute the PSI difference for each row (KO - Control)
compute_difference <- function(df) {
  avg_KO <- rowMeans(df[, 9:10])
  avg_control <- rowMeans(df[, 7:8])
  avg_KO - avg_control
}

# Calculate the difference and remove any additional NAs
fmndko_introns_df$difference <- compute_difference(fmndko_introns_df)
fmndko_introns_df <- na.omit(fmndko_introns_df)

# Create a group variable based on whether the difference is below or above zero
fmndko_introns_df$group <- ifelse(fmndko_introns_df$difference < 0, "Skipped", "Included")

# Perform a one-sample Wilcoxon signed-rank test (null hypothesis: median = 0)
symmetry_test <- wilcox.test(fmndko_introns_df$difference, mu = 0)

# Compute the percentages (areas under the density curve)
pct_below <- round(mean(fmndko_introns_df$difference < 0) * 100, 1)
pct_above <- round(mean(fmndko_introns_df$difference > 0) * 100, 1)

# ---- Plot -------------------------------------------------------------------

plot_psi_distribution <- ggplot(
  fmndko_introns_df[abs(fmndko_introns_df$difference) > 0,], 
  aes(y = ..density.., x = difference, fill = group)
) +
  geom_histogram(bins = 200, position = "identity", alpha = 0.9) +
  labs(
    title = "PSI Difference Distribution by Exon Group",
    subtitle = paste("One-sample Wilcoxon test p-value for Symmetry:", format(symmetry_test$p.value, digits = 3)),
    x = "ΔPSI (KO - Control)",
    y = "Density",
    fill = "Splicing Direction",
    caption = paste0("Created by AG on ", Sys.Date())  # Display quantiles in caption
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    legend.position = "right",
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.text.x = element_text(size = 15),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray80", size = 0.5),
    panel.grid.minor = element_line(color = "gray95", size = 0.3),
    strip.text = element_text(size = 14, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) +
  # Set the x-axis breaks
  scale_x_continuous(
    breaks = c(-100, -75, -50, -25, 0, 25, 50, 75, 100)
  ) +
  # Annotate the percentages on the left and right sides of the plot
  annotate("text", x=-65, y=0.1, hjust=0, vjust=0, label = paste0("dPSI<0: ",pct_below, "%"), color = "blue", size = 7) +
  annotate("text", x=65, y=0.1, hjust=1, vjust=0,
           label = paste0("dPSI>0: ",pct_above, "%"), color = "red", size = 7) +
  scale_fill_brewer(palette = "Set1")

print(plot_psi_distribution)
```




# PCA Plot { .tabset}


## SpireDKO

```{r pca calculation spire}
# Subset and scale data
pca_spire <- spire_data[, c("EVENT",spire_events$Samples)] %>%
  na.omit()

rownames(pca_spire)<- pca_spire$EVENT
pca_spire<-t(pca_spire[,-1])


pca_result_spire <- prcomp(pca_spire)  # Perform PCA

# Prepare PCA results for plotting
pca_data <- as.data.frame(pca_result_spire$x)
pca_data$Sample <- rownames(pca_data)
pca_data$condition<-metadata_spire$condition
```

```{r pca plotting spire}
# Create PCA plot
pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = as.factor(condition), )) +
  geom_point(size = 6) +
  labs(
    title = "<b style='font-size:18px;'>PCA of Samples (PSI Data)</b>",
    x = paste("PC1 (", round(100 * summary(pca_result_spire)$importance[2, 1], 1), "%)", sep = ""),
    y = paste("PC2 (", round(100 * summary(pca_result_spire)$importance[2, 2], 1), "%)", sep = ""),
    caption = paste0("Created by AG on ", Sys.Date()), size=3
  ) +
  theme_minimal(base_family = "roboto") +
  theme(
    plot.title = element_markdown(),
    plot.title.position = "plot",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "right",
    legend.title = element_blank(),
    panel.background = element_rect(fill = "gray98"),
    panel.grid = element_line(color = "gray90", size = 0.5),
    panel.grid.major = element_line(color = "gray90", size = 0.5),
    panel.grid.minor = element_blank()
  ) +
  scale_colour_paletteer_d("MetBrewer::Hokusai3") +
  coord_fixed()

pca_plot

```


## FMN2DKO

```{r pca calculation fmndko}
# Subset and scale data
pca_fmndko <- fmndko_data[, c("EVENT",fmndko_events$Samples)] %>%
  na.omit()

rownames(pca_fmndko)<- pca_fmndko$EVENT
pca_fmndko<-t(pca_fmndko[,-1])


pca_result_fmndko <- prcomp(pca_fmndko)  # Perform PCA

# Prepare PCA results for plotting
pca_data <- as.data.frame(pca_result_fmndko$x)
pca_data$Sample <- rownames(pca_data)
pca_data$condition<-metadata_fmn2dko$condition
```

```{r pca plotting fmndko}
# Create PCA plot
pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = as.factor(condition), )) +
  geom_point(size = 6) +
  labs(
    title = "<b style='font-size:18px;'>PCA of Samples (PSI Data)</b>",
    x = paste("PC1 (", round(100 * summary(pca_result_fmndko)$importance[2, 1], 1), "%)", sep = ""),
    y = paste("PC2 (", round(100 * summary(pca_result_fmndko)$importance[2, 2], 1), "%)", sep = ""),
    caption = paste0("Created by AG on ", Sys.Date()), size=3
  ) +
  theme_minimal(base_family = "roboto") +
  theme(
    plot.title = element_markdown(),
    plot.title.position = "plot",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "right",
    legend.title = element_blank(),
    panel.background = element_rect(fill = "gray98"),
    panel.grid = element_line(color = "gray90", size = 0.5),
    panel.grid.major = element_line(color = "gray90", size = 0.5),
    panel.grid.minor = element_blank()
  ) +
  scale_colour_paletteer_d("MetBrewer::Hokusai3") +
  coord_fixed()

pca_plot

```




## Exon Lists

```{r aux_featuregroup}
# Extract unique groups and sample IDs

# fmn2dko

metadata_fmn2dko<-data.frame(samples=c("SRR6026682_SRR6026683_SRR6026684_merged","SRR6026685_SRR6026686_SRR6026687_merged","SRR6026688_SRR6026689_SRR6026690_merged","SRR6026691_SRR6026692_SRR6026693_merged"),condition=c("control","control","fmndko","fmndko"))
groups_fmndko <- unique(metadata_fmn2dko[, "condition"])
samples_fmndko <- metadata_fmn2dko$samples

# spiredko

groups_spire <- unique(metadata_spire[, "condition"])
samples_spire <- metadata_spire$samples


random_colors=c("#D8D97AFF", "#95C36EFF", "#74C8C3FF", "#5A97C1FF", "#295384FF", "#0A2E57FF")


# Create group list with metadata
groupList_fmndko <- lapply(1:length(groups_fmndko), function(i) {
  list(
    name = groups_fmndko[i],
    samples = samples_fmndko[metadata_fmn2dko[, "condition"] == groups_fmndko[i]],
    color = random_colors[i]
  )
})
names(groupList_fmndko) <- groups_fmndko

# Create group list with metadata
groupList_spire <- lapply(1:length(groups_spire), function(i) {
  list(
    name = groups_spire[i],
    samples = samples_spire[metadata_spire[, "condition"] == groups_spire[i]],
    color = random_colors[i]
  )
})
names(groupList_spire) <- groups_spire

```

```{r groups_auxiliary}
# Define groups

groupA_fmndko<-"control"
groupB_fmndko<-"fmndko"

groupA_spire<-"control"
groupB_spire<-"spiredko"


samplesA_fmndko <- groupList_fmndko[[groupA_fmndko]]$samples
samplesB_fmndko <- groupList_fmndko[[groupB_fmndko]]$samples
colsGroupA_fmndko <- convertCols(fmndko_exons$PSI, samplesA_fmndko)
colsGroupB_fmndko <- convertCols(fmndko_exons$PSI, samplesB_fmndko)

samplesA_spire <- groupList_spire[[groupA_spire]]$samples
samplesB_spire <- groupList_spire[[groupB_spire]]$samples
colsGroupA_spire <- convertCols(spire_exons$PSI, samplesA_spire)
colsGroupB_spire <- convertCols(spire_exons$PSI, samplesB_spire)


set.seed(42) #Setting seed for downstream simulations of the beta distribution
```

### Calculations

```{r pdiff calculation exons}

# Prepare table for fmn2dko
fmndko_pdiff_exons <- prepareTableVolcanoFDR(
  psitable = fmndko_exons$PSI,
  qualtable = fmndko_exons$Qual,
  npoints = 500,
  colsA = colsGroupA_fmndko,
  colsB = colsGroupB_fmndko,
  labA = groupA_fmndko,
  labB = groupB_fmndko,
  basalColor = "#89C0AE",
  interestColor = "#E69A9C",
  maxDevTable = maxDevSimulationN100,
  nsim = 100,
  seed = TRUE,
  CoverageWeight = FALSE
)

# Prepare table for spire
spire_pdiff_exons <- prepareTableVolcanoFDR(
  psitable = spire_exons$PSI,
  qualtable = spire_exons$Qual,
  npoints = 500,
  colsA = colsGroupA_spire,
  colsB = colsGroupB_spire,
  labA = groupA_spire,
  labB = groupB_spire,
  basalColor = "#89C0AE",
  interestColor = "#E69A9C",
  maxDevTable = maxDevSimulationN100,
  nsim = 100,
  seed = TRUE,
  CoverageWeight = FALSE
)

differential_spire_exons<-na.omit(spire_pdiff_exons[spire_pdiff_exons$FDR <= 0.05 & abs(spire_pdiff_exons$deltapsi) >= 0.1,])
differential_fmndko_exons<-na.omit(fmndko_pdiff_exons[fmndko_pdiff_exons$FDR <= 0.05 & abs(fmndko_pdiff_exons$deltapsi) >= 0.1,])

shared_exons<-differential_spire_exons$EVENT[differential_spire_exons$EVENT %in% differential_fmndko_exons$EVENT]

write.csv(fmndko_pdiff_exons,"fmndko_pdiff_exons.csv")
write.csv(spire_pdiff_exons,"spire_pdiff_exons.csv")


```




### Volcano Plots { .tabset}


#### FMN2DKO

```{r fmndko fdr volcano exons}

library(ggrepel)
fmndko_pdiff_exons<-na.omit(fmndko_pdiff_exons)
# Add a Shared column based on the shared_exons
fmndko_pdiff_exons$Shared <- ifelse(fmndko_pdiff_exons$EVENT %in% shared_exons, "Shared", "Not Shared")

# Classify significance
fmndko_pdiff_exons$Significant <- ifelse(
  fmndko_pdiff_exons$FDR <=0.05 & fmndko_pdiff_exons$deltapsi >= 0.1, "Overrepresented",
  ifelse(
    fmndko_pdiff_exons$FDR <=0.05 & fmndko_pdiff_exons$deltapsi <= -0.1, "Underrepresented",
    "Not Significant"
  )
)

# Convert FDR to -log10 scale for y-axis
fmndko_pdiff_exons$negLogpvalue <- -log10(fmndko_pdiff_exons$FDR)


# Updated volcano plot with improved fonts and axis styling
volcano_plot <- ggplot(fmndko_pdiff_exons, aes(x = deltapsi, y = negLogpvalue)) +
  geom_point(aes(
    color = ifelse(Shared == "Shared", "Shared", Significant),
    alpha = ifelse(Shared == "Shared", "Shared", Significant),
    size = ifelse(Shared == "Shared", "Shared", Significant)
  )) +
  scale_color_manual(values = c(
    "Shared" = "yellow",
    "Overrepresented" = "#E00E73",
    "Underrepresented" = "#3E2672",
    "Not Significant" = "gray"
  )) +
  scale_alpha_manual(values = c(
    "Shared" = 1.0,  # Fully opaque for shared exons
    "Overrepresented" = 0.6,
    "Underrepresented" = 0.6,
    "Not Significant" = 0.3
  )) +
  scale_size_manual(values = c(
    "Shared" = 3,  # Larger size for shared exons
    "Overrepresented" = 2,
    "Underrepresented" = 2,
    "Not Significant" = 1
  )) +
  labs(
    x = "Delta PSI",
    y = expression(-log[10](FDR)),
    title = "Volcano Plot of FMN2DKO Exons",
    subtitle = "Shared events are highlighted in golden"
  ) +
  theme_minimal(base_size = 14, base_family = "Arial") +  # Set base font
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 18, face = "bold", color = "black"
    ),  # Centered and bold title
    plot.subtitle = element_text(
      hjust = 0.5, size = 14, face = "italic", color = "gray30"
    ),  # Italicized subtitle
    axis.text = element_text(size = 12, color = "black"),  # Axis text styling
    axis.title = element_text(size = 14, face = "bold", color = "black"),  # Axis title styling
    legend.text = element_text(size = 12),  # Larger legend text
    legend.title = element_blank(),  # Remove legend title
    panel.grid.major = element_line(color = "gray90"),  # Subtle grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.border = element_rect(color = "black", fill = NA, size = 1)  # Add border to plot area
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))  # Extend y-axis dynamically

# Display the plot
volcano_plot



```

#### Spire2

```{r spire fdr volcano exons}
spire_pdiff_exons<-na.omit(spire_pdiff_exons)
# Add a Shared column based on the shared_exons
spire_pdiff_exons$Shared <- ifelse(spire_pdiff_exons$EVENT %in% shared_exons, "Shared", "Not Shared")

# Classify significance
spire_pdiff_exons$Significant <- ifelse(
  spire_pdiff_exons$FDR <=0.05 & spire_pdiff_exons$deltapsi >= 0.1, "Overrepresented",
  ifelse(
    spire_pdiff_exons$FDR <=0.05 & spire_pdiff_exons$deltapsi <= -0.1, "Underrepresented",
    "Not Significant"
  )
)

# Convert FDR to -log10 scale for y-axis
spire_pdiff_exons$negLogpvalue <- -log10(spire_pdiff_exons$FDR)


# Updated volcano plot with improved fonts and axis styling
volcano_plot <- ggplot(spire_pdiff_exons, aes(x = deltapsi, y = negLogpvalue)) +
  geom_point(aes(
    color = ifelse(Shared == "Shared", "Shared", Significant),
    alpha = ifelse(Shared == "Shared", "Shared", Significant),
    size = ifelse(Shared == "Shared", "Shared", Significant)
  )) +
  scale_color_manual(values = c(
    "Shared" = "yellow",
    "Overrepresented" = "#E00E73",
    "Underrepresented" = "#3E2672",
    "Not Significant" = "gray"
  )) +
  scale_alpha_manual(values = c(
    "Shared" = 1.0,  # Fully opaque for shared exons
    "Overrepresented" = 0.6,
    "Underrepresented" = 0.6,
    "Not Significant" = 0.3
  )) +
  scale_size_manual(values = c(
    "Shared" = 3,  # Larger size for shared exons
    "Overrepresented" = 2,
    "Underrepresented" = 2,
    "Not Significant" = 1
  )) +
  labs(
    x = "Delta PSI",
    y = expression(-log[10](FDR)),
    title = "Volcano Plot of Spire Exons",
    subtitle = "Shared events are highlighted in golden"
  ) +
  theme_minimal(base_size = 14, base_family = "Arial") +  # Set base font
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 18, face = "bold", color = "black"
    ),  # Centered and bold title
    plot.subtitle = element_text(
      hjust = 0.5, size = 14, face = "italic", color = "gray30"
    ),  # Italicized subtitle
    axis.text = element_text(size = 12, color = "black"),  # Axis text styling
    axis.title = element_text(size = 14, face = "bold", color = "black"),  # Axis title styling
    legend.text = element_text(size = 12),  # Larger legend text
    legend.title = element_blank(),  # Remove legend title
    panel.grid.major = element_line(color = "gray90"),  # Subtle grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.border = element_rect(color = "black", fill = NA, size = 1)  # Add border to plot area
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))  # Extend y-axis dynamically

# Display the plot
volcano_plot


```


### Individual Combination Tables { .tabset}

These tables show the exons that show FDR <= 0.1 in the pairwise comparison of each condition. 


#### FMN2DKO

```{r fmndko combination table exons}
# Render DataTable with enhancements
datatable(
  differential_fmndko_exons,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "FDR",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)


```

#### Spire2


```{r spire combination table exons}
# Render DataTable with enhancements
datatable(
  differential_spire_exons,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "FDR",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)


```



## Intron Lists

### Calculations

```{r pdiff calculation introns}


# Prepare table for Tao
fmndko_pdiff_introns <- prepareTableVolcanoFDR(
  psitable = fmndko_introns$PSI,
  qualtable = fmndko_introns$Qual,
  npoints = 500,
  colsA = colsGroupA_fmndko,
  colsB = colsGroupB_fmndko,
  labA = groupA_fmndko,
  labB = groupB_fmndko,
  basalColor = "#89C0AE",
  interestColor = "#E69A9C",
  maxDevTable = maxDevSimulationN100,
  nsim = 100,
  seed = TRUE,
  CoverageWeight = FALSE
)

# Prepare table for spire
spire_pdiff_introns <- prepareTableVolcanoFDR(
  psitable = spire_introns$PSI,
  qualtable = spire_introns$Qual,
  npoints = 500,
  colsA = colsGroupA_spire,
  colsB = colsGroupB_spire,
  labA = groupA_spire,
  labB = groupB_spire,
  basalColor = "#89C0AE",
  interestColor = "#E69A9C",
  maxDevTable = maxDevSimulationN100,
  nsim=100,
  seed = TRUE,
  CoverageWeight = FALSE
)

differential_spire_introns<-na.omit(spire_pdiff_introns[spire_pdiff_introns$FDR <=0.05 & abs(spire_pdiff_introns$deltapsi)>=0.1,])
differential_fmndko_introns<-na.omit(fmndko_pdiff_introns[fmndko_pdiff_introns$FDR <=0.05 & abs(fmndko_pdiff_introns$deltapsi)>=0.1,])



shared_introns<-differential_spire_introns$EVENT[differential_spire_introns$EVENT %in% differential_fmndko_introns$EVENT]

write.csv(fmndko_pdiff_introns,"fmndko_pdiff_introns.csv")
write.csv(spire_pdiff_introns,"spire_pdiff_introns.csv")

```



### Volcano Plots { .tabset}



```{r fmndko volcano introns}
fmndko_pdiff_introns<-na.omit(fmndko_pdiff_introns)
# Add a Shared column based on the shared_exons
fmndko_pdiff_introns$Shared <- ifelse(fmndko_pdiff_introns$EVENT %in% shared_introns, "Shared", "Not Shared")

# Classify significance
fmndko_pdiff_introns$Significant <- ifelse(
  fmndko_pdiff_introns$FDR <=0.05 & fmndko_pdiff_introns$deltapsi >= 0.1, "Overrepresented",
  ifelse(
    fmndko_pdiff_introns$FDR <=0.05 & fmndko_pdiff_introns$deltapsi <= -0.1, "Underrepresented",
    "Not Significant"
  )
)

# Convert FDR to -log10 scale for y-axis
fmndko_pdiff_introns$negLogpvalue <- -log10(fmndko_pdiff_introns$FDR)


# Updated volcano plot with improved fonts and axis styling
volcano_plot <- ggplot(fmndko_pdiff_introns, aes(x = deltapsi, y = negLogpvalue)) +
  geom_point(aes(
    color = ifelse(Shared == "Shared", "Shared", Significant),
    alpha = ifelse(Shared == "Shared", "Shared", Significant),
    size = ifelse(Shared == "Shared", "Shared", Significant)
  )) +
  scale_color_manual(values = c(
    "Shared" = "yellow",
    "Overrepresented" = "#E00E73",
    "Underrepresented" = "#3E2672",
    "Not Significant" = "gray"
  )) +
  scale_alpha_manual(values = c(
    "Shared" = 1.0,  # Fully opaque for shared exons
    "Overrepresented" = 0.6,
    "Underrepresented" = 0.6,
    "Not Significant" = 0.3
  )) +
  scale_size_manual(values = c(
    "Shared" = 3,  # Larger size for shared exons
    "Overrepresented" = 2,
    "Underrepresented" = 2,
    "Not Significant" = 1
  )) +
  labs(
    x = "Delta PSI",
    y = expression(-log[10](FDR)),
    title = "Volcano Plot of FMN2DKO Introns",
    subtitle = "Shared events are highlighted in golden"
  ) +
  theme_minimal(base_size = 14, base_family = "Arial") +  # Set base font
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 18, face = "bold", color = "black"
    ),  # Centered and bold title
    plot.subtitle = element_text(
      hjust = 0.5, size = 14, face = "italic", color = "gray30"
    ),  # Italicized subtitle
    axis.text = element_text(size = 12, color = "black"),  # Axis text styling
    axis.title = element_text(size = 14, face = "bold", color = "black"),  # Axis title styling
    legend.text = element_text(size = 12),  # Larger legend text
    legend.title = element_blank(),  # Remove legend title
    panel.grid.major = element_line(color = "gray90"),  # Subtle grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.border = element_rect(color = "black", fill = NA, size = 1)  # Add border to plot area
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))  # Extend y-axis dynamically

# Display the plot
volcano_plot




```

#### Spire2

```{r spire volcano introns}
spire_pdiff_introns<-na.omit(spire_pdiff_introns)
spire_pdiff_introns$Shared <- ifelse(spire_pdiff_introns$EVENT %in% shared_introns, "Shared", "Not Shared")

# Classify significance
spire_pdiff_introns$Significant <- ifelse(
  spire_pdiff_introns$FDR <=0.05 & spire_pdiff_introns$deltapsi >= 0.1, "Overrepresented",
  ifelse(
    spire_pdiff_introns$FDR <=0.05 & spire_pdiff_introns$deltapsi <= -0.1, "Underrepresented",
    "Not Significant"
  )
)

# Convert FDR to -log10 scale for y-axis
spire_pdiff_introns$negLogpvalue <- -log10(spire_pdiff_introns$FDR)


# Updated volcano plot with improved fonts and axis styling
volcano_plot <- ggplot(spire_pdiff_introns, aes(x = deltapsi, y = negLogpvalue)) +
  geom_point(aes(
    color = ifelse(Shared == "Shared", "Shared", Significant),
    alpha = ifelse(Shared == "Shared", "Shared", Significant),
    size = ifelse(Shared == "Shared", "Shared", Significant)
  )) +
  scale_color_manual(values = c(
    "Shared" = "yellow",
    "Overrepresented" = "#E00E73",
    "Underrepresented" = "#3E2672",
    "Not Significant" = "gray"
  )) +
  scale_alpha_manual(values = c(
    "Shared" = 1.0,  # Fully opaque for shared exons
    "Overrepresented" = 0.6,
    "Underrepresented" = 0.6,
    "Not Significant" = 0.3
  )) +
  scale_size_manual(values = c(
    "Shared" = 3,  # Larger size for shared exons
    "Overrepresented" = 2,
    "Underrepresented" = 2,
    "Not Significant" = 1
  )) +
  labs(
    x = "Delta PSI",
    y = expression(-log[10](FDR)),
    title = "Volcano Plot of Spire Introns",
    subtitle = "Shared events are highlighted in golden"
  ) +
  theme_minimal(base_size = 14, base_family = "Arial") +  # Set base font
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 18, face = "bold", color = "black"
    ),  # Centered and bold title
    plot.subtitle = element_text(
      hjust = 0.5, size = 14, face = "italic", color = "gray30"
    ),  # Italicized subtitle
    axis.text = element_text(size = 12, color = "black"),  # Axis text styling
    axis.title = element_text(size = 14, face = "bold", color = "black"),  # Axis title styling
    legend.text = element_text(size = 12),  # Larger legend text
    legend.title = element_blank(),  # Remove legend title
    panel.grid.major = element_line(color = "gray90"),  # Subtle grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.border = element_rect(color = "black", fill = NA, size = 1)  # Add border to plot area
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))  # Extend y-axis dynamically

# Display the plot
volcano_plot




```

### Individual Combination Tables { .tabset}

These tables show the introns that show FDR <= 0.1 in the pairwise comparison of each condition. 

#### FMN2DKO



```{r fmndko combination table introns}
# Render DataTable with enhancements
datatable(
  differential_fmndko_introns,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "FDR",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)


```

#### Spire2


```{r spire combination table introns}
# Render DataTable with enhancements
datatable(
  differential_spire_introns,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "FDR",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)

```


## Alternative Splicing List

### Calculations

```{r pdiff calculation alt}

# Prepare table for Tao
fmndko_pdiff_alt <- prepareTableVolcanoFDR(
  psitable = fmndko_alt$PSI,
  qualtable = fmndko_alt$Qual,
  npoints = 500,
  colsA = colsGroupA_fmndko,
  colsB = colsGroupB_fmndko,
  labA = groupA_fmndko,
  labB = groupB_fmndko,
  basalColor = "#89C0AE",
  interestColor = "#E69A9C",
  maxDevTable = maxDevSimulationN100,
  nsim=100,
  seed = TRUE,
  CoverageWeight = FALSE
)

# Prepare table for spire
spire_pdiff_alt <- prepareTableVolcanoFDR(
  psitable = spire_alt$PSI,
  qualtable = spire_alt$Qual,
  npoints = 500,
  colsA = colsGroupA_spire,
  colsB = colsGroupB_spire,
  labA = groupA_spire,
  labB = groupB_spire,
  basalColor = "#89C0AE",
  interestColor = "#E69A9C",
  maxDevTable = maxDevSimulationN100,
  nsim=100,
  seed = TRUE,
  CoverageWeight = FALSE
)

differential_spire_alt<-na.omit(spire_pdiff_alt[spire_pdiff_alt$FDR <=0.05 & abs(spire_pdiff_alt$deltapsi)>=0.1 ,])
differential_fmndko_alt<-na.omit(fmndko_pdiff_alt[fmndko_pdiff_alt$FDR <=0.05 & abs(fmndko_pdiff_alt$deltapsi)>=0.1 ,])

shared_alt<-differential_spire_alt$EVENT[differential_spire_alt$EVENT %in% differential_fmndko_alt$EVENT]

write.csv(fmndko_pdiff_alt,"fmndko_pdiff_alt")
write.csv(spire_pdiff_alt,"spire_pdiff_alt.csv")
```




### Volcano Plots { .tabset}


#### FMN2DKO

```{r fmndko  volcano alt}
fmndko_pdiff_alt<-na.omit(fmndko_pdiff_alt)
fmndko_pdiff_alt$Shared <- ifelse(fmndko_pdiff_alt$EVENT %in% shared_alt, "Shared", "Not Shared")

# Classify significance
fmndko_pdiff_alt$Significant <- ifelse(
  fmndko_pdiff_alt$FDR <=0.05 & fmndko_pdiff_alt$deltapsi >= 0.1, "Overrepresented",
  ifelse(
    fmndko_pdiff_alt$FDR <=0.05 & fmndko_pdiff_alt$deltapsi <= -0.1, "Underrepresented",
    "Not Significant"
  )
)

# Convert FDR to -log10 scale for y-axis
fmndko_pdiff_alt$negLogpvalue <- -log10(fmndko_pdiff_alt$FDR)


# Updated volcano plot with improved fonts and axis styling
volcano_plot <- ggplot(fmndko_pdiff_alt, aes(x = deltapsi, y = negLogpvalue)) +
  geom_point(aes(
    color = ifelse(Shared == "Shared", "Shared", Significant),
    alpha = ifelse(Shared == "Shared", "Shared", Significant),
    size = ifelse(Shared == "Shared", "Shared", Significant)
  )) +
  scale_color_manual(values = c(
    "Shared" = "yellow",
    "Overrepresented" = "#E00E73",
    "Underrepresented" = "#3E2672",
    "Not Significant" = "gray"
  )) +
  scale_alpha_manual(values = c(
    "Shared" = 1.0,  # Fully opaque for shared exons
    "Overrepresented" = 0.6,
    "Underrepresented" = 0.6,
    "Not Significant" = 0.3
  )) +
  scale_size_manual(values = c(
    "Shared" = 3,  # Larger size for shared exons
    "Overrepresented" = 2,
    "Underrepresented" = 2,
    "Not Significant" = 1
  )) +
  labs(
    x = "Delta PSI",
    y = expression(-log[10](FDR)),
    title = "Volcano Plot of FMN2DKO Alternative 5´/3´",
    subtitle = "Shared events are highlighted in golden"
  ) +
  theme_minimal(base_size = 14, base_family = "Arial") +  # Set base font
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 18, face = "bold", color = "black"
    ),  # Centered and bold title
    plot.subtitle = element_text(
      hjust = 0.5, size = 14, face = "italic", color = "gray30"
    ),  # Italicized subtitle
    axis.text = element_text(size = 12, color = "black"),  # Axis text styling
    axis.title = element_text(size = 14, face = "bold", color = "black"),  # Axis title styling
    legend.text = element_text(size = 12),  # Larger legend text
    legend.title = element_blank(),  # Remove legend title
    panel.grid.major = element_line(color = "gray90"),  # Subtle grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.border = element_rect(color = "black", fill = NA, size = 1)  # Add border to plot area
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))  # Extend y-axis dynamically

# Display the plot
volcano_plot



```

#### Spire2

```{r spire fdr volcano alt}
spire_pdiff_alt<-na.omit(spire_pdiff_alt)
spire_pdiff_alt$Shared <- ifelse(spire_pdiff_alt$EVENT %in% shared_alt, "Shared", "Not Shared")

# Classify significance
spire_pdiff_alt$Significant <- ifelse(
  spire_pdiff_alt$FDR <=0.05 & spire_pdiff_alt$deltapsi >= 0.1, "Overrepresented",
  ifelse(
    spire_pdiff_alt$FDR <=0.05 & spire_pdiff_alt$deltapsi <= -0.1, "Underrepresented",
    "Not Significant"
  )
)

# Convert FDR to -log10 scale for y-axis
spire_pdiff_alt$negLogpvalue <- -log10(spire_pdiff_alt$FDR)


# Updated volcano plot with improved fonts and axis styling
volcano_plot <- ggplot(spire_pdiff_alt, aes(x = deltapsi, y = negLogpvalue)) +
  geom_point(aes(
    color = ifelse(Shared == "Shared", "Shared", Significant),
    alpha = ifelse(Shared == "Shared", "Shared", Significant),
    size = ifelse(Shared == "Shared", "Shared", Significant)
  )) +
  scale_color_manual(values = c(
    "Shared" = "yellow",
    "Overrepresented" = "#E00E73",
    "Underrepresented" = "#3E2672",
    "Not Significant" = "gray"
  )) +
  scale_alpha_manual(values = c(
    "Shared" = 1.0,  # Fully opaque for shared exons
    "Overrepresented" = 0.6,
    "Underrepresented" = 0.6,
    "Not Significant" = 0.3
  )) +
  scale_size_manual(values = c(
    "Shared" = 4,  # Larger size for shared exons
    "Overrepresented" = 2,
    "Underrepresented" = 2,
    "Not Significant" = 1
  )) +
  labs(
    x = "Delta PSI",
    y = expression(-log[10](FDR)),
    title = "Volcano Plot of Spire Alternative 5´/3´",
    subtitle = "Shared events are highlighted in golden"
  ) +
  theme_minimal(base_size = 14, base_family = "Arial") +  # Set base font
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 18, face = "bold", color = "black"
    ),  # Centered and bold title
    plot.subtitle = element_text(
      hjust = 0.5, size = 14, face = "italic", color = "gray30"
    ),  # Italicized subtitle
    axis.text = element_text(size = 12, color = "black"),  # Axis text styling
    axis.title = element_text(size = 14, face = "bold", color = "black"),  # Axis title styling
    legend.text = element_text(size = 12),  # Larger legend text
    legend.title = element_blank(),  # Remove legend title
    panel.grid.major = element_line(color = "gray90"),  # Subtle grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.border = element_rect(color = "black", fill = NA, size = 1)  # Add border to plot area
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))  # Extend y-axis dynamically

# Display the plot
volcano_plot



```

### Individual Combination Tables { .tabset}

These tables show the Alt5 or Alt3 that show FDR <= 0.1 in the pairwise comparison of each condition. 

#### FMN2DKO


```{r fmndko combination table alt}
# Render DataTable with enhancements
datatable(
  differential_fmndko_alt,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "FDR",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)


```

#### Spire2



```{r spire combination table alt}
# Render DataTable with enhancements
datatable(
  differential_spire_alt,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "FDR",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)


```

## Comparing both Analysis

The following table compares the results of the two analyses: the one using the complete dataset (this report) or the simpler one, only using alternative events. The more complete analysis finds more significant (FDR <= 0.05 & |dPSI| >= 0.1) splicing events. It also preserves the majority of those found by the simpler analysis. 


```{r comparison}
old_spire_exons<-filter(read.csv("versionA_lists/exon_spire.csv")[,-1], FDR<=0.05, abs(deltapsi)>=0.1)
old_fmndko_exons<-filter(read.csv("versionA_lists/exon_fmndko.csv")[,-1], FDR<=0.05, abs(deltapsi)>=0.1)
old_spire_introns<-filter(read.csv("versionA_lists/intron_spire.csv")[,-1], FDR<=0.05, abs(deltapsi)>=0.1)
old_fmndko_introns<-filter(read.csv("versionA_lists/intron_fmndko.csv")[,-1], FDR<=0.05, abs(deltapsi)>=0.1)
old_spire_alt<-filter(read.csv("versionA_lists/alt_spire.csv")[,-1], FDR<=0.05, abs(deltapsi)>=0.1)
old_fmndko_alt<-filter(read.csv("versionA_lists/alt_fmndko.csv")[,-1], FDR<=0.05, abs(deltapsi)>=0.1)

compare_table<-data.frame(study=c("FMN2 (alternative/complete)", "Spire (alternative/complete)"), number_exons=c(paste0(length(unique(old_fmndko_exons$EVENT)),"/",length(unique(differential_fmndko_exons$EVENT))), paste0(length(unique(old_spire_exons$EVENT)),"/",length(unique(differential_spire_exons$EVENT)))), number_introns=c(paste0(length(unique(old_fmndko_introns$EVENT)),"/",length(unique(differential_fmndko_introns$EVENT))), paste0(length(unique(old_spire_introns$EVENT)),"/",length(unique(differential_spire_introns$EVENT)))), number_alt=c(paste0(length(unique(old_fmndko_alt$EVENT)),"/",length(unique(differential_fmndko_alt$EVENT))), paste0(length(unique(old_spire_alt$EVENT)),"/",length(unique(differential_spire_alt$EVENT)))), overlapping=c(paste0(round(mean(c(old_fmndko_exons$EVENT,old_fmndko_introns$EVENT,old_fmndko_alt$EVENT) %in% c(differential_fmndko_exons$EVENT,differential_fmndko_introns$EVENT,differential_fmndko_alt$EVENT))*100, digits = 1), "%"), paste0(round(mean(c(old_spire_exons$EVENT,old_spire_introns$EVENT,old_spire_alt$EVENT) %in% c(differential_spire_exons$EVENT,differential_spire_introns$EVENT,differential_spire_alt$EVENT))*100, digits = 1), "%")))


compare_table <- compare_table %>%
  rename(
    "Significant Exons" = number_exons,
    "Significant Introns" = number_introns,
    "Significant Alt 5´ 3´" = number_alt,
    "Overlapping"=overlapping
  )

datatable(compare_table)
```


## Enrichment Analysis with EnrichR { .tabset}


### FMN2DKO { .tabset}

```{r fmndko enrichr}

enrichdata<-enrichr(unique(c(differential_fmndko_exons$GENE, differential_fmndko_introns$GENE, differential_fmndko_alt$GENE,)), databases = c("GO_Biological_Process_2023","GO_Cellular_Component_2023","GO_Molecular_Function_2023"))


```


#### GO Biological Process

```{r fmndko bioprocess, fig.width=10, fig.height=8}
plotEnrich(enrichdata[[1]])

datatable(
  enrichdata[[1]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Cellular Component

```{r fmndko cellcomponent, fig.width=10, fig.height=8}
plotEnrich(enrichdata[[2]])

datatable(
  enrichdata[[2]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Molecular Function

```{r fmndko molfunction, fig.width=10, fig.height=8}
plotEnrich(enrichdata[[3]])
datatable(
  enrichdata[[3]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

### Spire { .tabset}

```{r spire enrichr}
enrichdata<-enrichr(unique(c(differential_spire_exons$GENE, differential_spire_introns$GENE, differential_spire_alt$GENE)), databases = c("GO_Biological_Process_2023","GO_Cellular_Component_2023","GO_Molecular_Function_2023"))


```

#### GO Biological Process

```{r spire bioprocess, fig.width=10, fig.height=8}
plotEnrich(enrichdata[[1]])
datatable(
  enrichdata[[1]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Cellular Component

```{r spire cellcomponent, fig.width=10, fig.height=8}
plotEnrich(enrichdata[[2]])
datatable(
  enrichdata[[2]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Molecular Function

```{r spire molfunction, fig.width=10, fig.height=8}
plotEnrich(enrichdata[[3]])
datatable(
  enrichdata[[3]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```


# Gene-centric Analysis { .tabset}


## FMNDKO

```{r gene-centric plot}
set.seed(44)
protein_impact <- read.table("PROT_IMPACT-mm10-v2.3.tab", sep = "\t",
                             header = TRUE, stringsAsFactors = FALSE, fill = TRUE, quote = "")
# -------------------------------
# 1. Combine data and add protein impact
# -------------------------------
combined_df <- bind_rows(
  exons  = differential_fmndko_exons,
  introns   = differential_fmndko_introns,
  .id = "source"   # new column indicating the origin
) 

# Match the protein impact values
combined_df$protein_impact <- protein_impact$ONTO[match(combined_df$EVENT, protein_impact$EventID)]

# select only unqiue values in GENE
combined_df <- combined_df %>% distinct(EVENT, .keep_all = TRUE)
# -------------------------------
# 2. Create per-gene summary data
# -------------------------------
combined_df_summary <- combined_df %>%
  distinct(GENE, EVENT, .keep_all = TRUE) %>%  # remove duplicate gene/event rows if any
  group_by(GENE) %>%
  summarise(
    nEvents   = n(),                                        # number of events per gene
    meanDelta = mean(abs(deltapsi), na.rm = TRUE),          # mean absolute deltapsi
    sdDelta   = sd(abs(deltapsi), na.rm = TRUE)             # standard deviation (for reference)
  ) %>%
  arrange(desc(nEvents), desc(meanDelta)) %>%              # sort so that genes with more events come first
  mutate(GENE = factor(GENE, levels = unique(GENE)))        # set the ordering for the x-axis

# -------------------------------
# 3. Update original data and join summary info
# -------------------------------
combined_df <- combined_df %>%
  mutate(
    absDelta = abs(deltapsi),
    impact   = ifelse(grepl("ORF", protein_impact, ignore.case = TRUE), "ORF", "Non-ORF")
  ) %>%
  left_join(combined_df_summary %>% select(GENE, nEvents), by = "GENE") %>%
  mutate(GENE = factor(GENE, levels = levels(combined_df_summary$GENE)))

# -------------------------------
# 4. Create custom facet strip labels (optional)
# -------------------------------
facet_labels <- combined_df_summary %>%
  group_by(nEvents) %>%
  summarise(n_genes = n()) %>%
  mutate(label = paste0(nEvents, " events (", n_genes, " genes)")) %>%
  deframe()  # a named vector mapping nEvents -> "n events (x genes)"

# -------------------------------
# 5. Create a mapping for x-axis labels: blank out gene names for nEvents==1 or 2
# -------------------------------
gene_labels <- combined_df_summary %>%
  mutate(gene_label = ifelse(nEvents %in% c(1, 2), "", as.character(GENE))) %>%
  select(GENE, gene_label) %>%
  deframe()

# -------------------------------
# 6. Create annotation data for facets with nEvents 1 and 2
# -------------------------------
annotation_df <- combined_df_summary %>%
  filter(nEvents %in% c(1, 2)) %>%
  group_by(nEvents) %>%
  summarise(
    n_genes      = n(),
    max_meanDelta= max(meanDelta)
  ) %>%
  # Here we compute an approximate x-location (center of the facet)
  # Note: In a free_x facet the bars are placed at positions 1,2,..., so (n_genes+1)/2 is roughly centered.
  mutate(
    x = (n_genes + 1) / 2,
    y = max_meanDelta * 1.05  # position the label slightly above the tallest bar in that facet
  )

# -------------------------------
# 7. Build the plot
# -------------------------------
p <- ggplot() +
  # Bar plot: summary values (mean absolute deltapsi) per gene
  geom_bar(
    data = combined_df_summary[-which(combined_df_summary$GENE==""),], 
    aes(x = GENE, y = meanDelta, fill = factor(nEvents)),
    stat = "identity", 
    width = 0.7, 
    alpha = 0.8
  ) +
  # Jitter: individual event values, with shape by protein impact
  geom_jitter(
    data = combined_df[-which(combined_df$GENE==""),],
    aes(x = GENE, y = absDelta, shape = impact),
    width = 0.15,    # small horizontal jitter
    size = 2.5, 
    alpha = 0.6,
    color = "black"
  ) +
  # Add a text annotation in facets where nEvents==1 or 2 showing the number of genes
  geom_text(
    data = annotation_df,
    aes(x = x, y = y, label = paste(n_genes, "genes")),
    inherit.aes = FALSE,
    size = 5,
    color = "black"
  ) +
  # Use a color palette for the bars and manually define the shapes
  scale_fill_brewer(palette = "Set2", name = "Number of Events") +
  scale_shape_manual(values = c("Non-ORF" = 16, "ORF" = 17), name = "Protein Impact") +
  # Apply the custom x-axis labels (gene names only appear for genes with >2 events)
  scale_x_discrete(labels = gene_labels) +
  labs(
    x = "Gene ID",
    y = "|dPSI|",
    title="FMNDKO",
    subtitle = "Mean |dPSI| of genes grouped by the number significant Splicing Events"
  ) +
  theme_bw(base_size = 14) +
  theme(
    # Rotate x-axis labels (for facets that show them)
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title   = element_text(hjust = 0.5),
    # Improved facet strip aesthetics:
    strip.background = element_rect(fill = "lightblue", color = "black"),
    strip.text       = element_text(face = "bold"),
    panel.background = element_blank(),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  # Facet the plot by the number of events (with free x-axis scales)
  facet_wrap(~ nEvents, scales = "free_x", labeller = labeller(nEvents = facet_labels))

# Display the plot
print(p)


```

## Spire

```{r gene-centric plot spire}
set.seed(44)
protein_impact <- read.table("PROT_IMPACT-mm10-v2.3.tab", sep = "\t",
                             header = TRUE, stringsAsFactors = FALSE, fill = TRUE, quote = "")
# -------------------------------
# 1. Combine data and add protein impact
# -------------------------------
combined_df <- bind_rows(
  exons  = differential_spire_exons,
  introns   = differential_spire_introns,
  .id = "source"   # new column indicating the origin
) 

# Match the protein impact values
combined_df$protein_impact <- protein_impact$ONTO[match(combined_df$EVENT, protein_impact$EventID)]

# select only unqiue values in GENE
combined_df <- combined_df %>% distinct(EVENT, .keep_all = TRUE)
# -------------------------------
# 2. Create per-gene summary data
# -------------------------------
combined_df_summary <- combined_df %>%
  distinct(GENE, EVENT, .keep_all = TRUE) %>%  # remove duplicate gene/event rows if any
  group_by(GENE) %>%
  summarise(
    nEvents   = n(),                                        # number of events per gene
    meanDelta = mean(abs(deltapsi), na.rm = TRUE),          # mean absolute deltapsi
    sdDelta   = sd(abs(deltapsi), na.rm = TRUE)             # standard deviation (for reference)
  ) %>%
  arrange(desc(nEvents), desc(meanDelta)) %>%              # sort so that genes with more events come first
  mutate(GENE = factor(GENE, levels = unique(GENE)))        # set the ordering for the x-axis

# -------------------------------
# 3. Update original data and join summary info
# -------------------------------
combined_df <- combined_df %>%
  mutate(
    absDelta = abs(deltapsi),
    impact   = ifelse(grepl("ORF", protein_impact, ignore.case = TRUE), "ORF", "Non-ORF")
  ) %>%
  left_join(combined_df_summary %>% select(GENE, nEvents), by = "GENE") %>%
  mutate(GENE = factor(GENE, levels = levels(combined_df_summary$GENE)))

# -------------------------------
# 4. Create custom facet strip labels (optional)
# -------------------------------
facet_labels <- combined_df_summary %>%
  group_by(nEvents) %>%
  summarise(n_genes = n()) %>%
  mutate(label = paste0(nEvents, " events (", n_genes, " genes)")) %>%
  deframe()  # a named vector mapping nEvents -> "n events (x genes)"

# -------------------------------
# 5. Create a mapping for x-axis labels: blank out gene names for nEvents==1 or 2
# -------------------------------
gene_labels <- combined_df_summary %>%
  mutate(gene_label = ifelse(nEvents %in% c(1, 2), "", as.character(GENE))) %>%
  select(GENE, gene_label) %>%
  deframe()

# -------------------------------
# 6. Create annotation data for facets with nEvents 1 and 2
# -------------------------------
annotation_df <- combined_df_summary %>%
  filter(nEvents %in% c(1, 2)) %>%
  group_by(nEvents) %>%
  summarise(
    n_genes      = n(),
    max_meanDelta= max(meanDelta)
  ) %>%
  # Here we compute an approximate x-location (center of the facet)
  # Note: In a free_x facet the bars are placed at positions 1,2,..., so (n_genes+1)/2 is roughly centered.
  mutate(
    x = (n_genes + 1) / 2,
    y = max_meanDelta * 1.05  # position the label slightly above the tallest bar in that facet
  )

# -------------------------------
# 7. Build the plot
# -------------------------------
p <- ggplot() +
  # Bar plot: summary values (mean absolute deltapsi) per gene
  geom_bar(
    data = combined_df_summary[-which(combined_df_summary$GENE==""),], 
    aes(x = GENE, y = meanDelta, fill = factor(nEvents)),
    stat = "identity", 
    width = 0.7, 
    alpha = 0.8
  ) +
  # Jitter: individual event values, with shape by protein impact
  geom_jitter(
    data = combined_df[-which(combined_df$GENE==""),],
    aes(x = GENE, y = absDelta, shape = impact),
    width = 0.15,    # small horizontal jitter
    size = 2.5, 
    alpha = 0.6,
    color = "black"
  ) +
  # Add a text annotation in facets where nEvents==1 or 2 showing the number of genes
  geom_text(
    data = annotation_df,
    aes(x = x, y = y, label = paste(n_genes, "genes")),
    inherit.aes = FALSE,
    size = 5,
    color = "black"
  ) +
  # Use a color palette for the bars and manually define the shapes
  scale_fill_brewer(palette = "Set2", name = "Number of Events") +
  scale_shape_manual(values = c("Non-ORF" = 16, "ORF" = 17), name = "Protein Impact") +
  # Apply the custom x-axis labels (gene names only appear for genes with >2 events)
  scale_x_discrete(labels = gene_labels) +
  labs(
    x = "Gene ID",
    y = "|dPSI|",
    title="Spire",
    subtitle = "Mean |dPSI| of genes grouped by the number significant Splicing Events"
  ) +
  theme_bw(base_size = 14) +
  theme(
    # Rotate x-axis labels (for facets that show them)
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title   = element_text(hjust = 0.5),
    # Improved facet strip aesthetics:
    strip.background = element_rect(fill = "lightblue", color = "black"),
    strip.text       = element_text(face = "bold"),
    panel.background = element_blank(),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  # Facet the plot by the number of events (with free x-axis scales)
  facet_wrap(~ nEvents, scales = "free_x", labeller = labeller(nEvents = facet_labels))

# Display the plot
print(p)


```

# Spliceosome Ontology Genes { .tabset}


## Spire

```{r, splicesosomo ontology behaviour}

enrichdata<-enrichr(unique(c(differential_spire_exons$GENE, differential_spire_introns$GENE, differential_spire_alt$GENE,differential_fmndko_exons$GENE, differential_fmndko_introns$GENE, differential_fmndko_alt$GENE)), databases = c("GO_Biological_Process_2023","GO_Cellular_Component_2023","GO_Molecular_Function_2023"))

# Extract and format gene names
genes_splicing_1 <- enrichdata[[1]] %>%
  filter(str_detect(Term, "Splic")) %>%  # Select rows with 'splic' in Term
  pull(Genes) %>%                         # Extract Genes column
  str_split(";") %>%                      # Split gene names
  unlist() %>%                            # Flatten list
  str_to_lower() %>%                      # Convert to lowercase
  str_replace("^.", toupper) 
# Capitalize first letter
genes_splicing_2 <- enrichdata[[2]] %>%
  filter(str_detect(Term, "splic") | str_detect(Term, "Splic")) %>%  # Select rows with 'splic' in Term
  pull(Genes) %>%                         # Extract Genes column
  str_split(";") %>%                      # Split gene names
  unlist() %>%                            # Flatten list
  str_to_lower() %>%                      # Convert to lowercase
  str_replace("^.", toupper) 

genes_splicing<-unique(c(genes_splicing_1,genes_splicing_2))

combined_df <- bind_rows(
  exons_spire = differential_spire_exons,
  introns_spire = differential_spire_introns,
  .id = "source"  # This will be the new column indicating the origin
)
splicing_df<-filter(combined_df, GENE %in% genes_splicing)
splicing_df$protein_impact<-protein_impact$ONTO[match(splicing_df$EVENT, protein_impact$EventID)]



# Prepare the summary data frame, ensuring proper sorting and factor ordering
splicing_df_summary <- splicing_df %>%
  # Remove duplicate rows for the same gene and event (if applicable)
  distinct(GENE, EVENT, .keep_all = TRUE) %>%
  group_by(GENE) %>%
  summarise(
    nEvents   = n(),                                        # count unique events per gene
    meanDelta = mean(abs(deltapsi), na.rm = TRUE),          # mean of absolute deltapsi
    sdDelta   = sd(abs(deltapsi), na.rm = TRUE)             # standard deviation (for reference)
  ) %>%
  # First sort by the number of events (desc), then by mean absolute deltapsi (desc)
  arrange(desc(nEvents), desc(meanDelta)) %>%
  # Set the factor levels for GENE according to the sorted order
  mutate(GENE = factor(GENE, levels = unique(GENE)))

# Update the original data frame:
# - Create a column for the absolute value of deltapsi.
# - Ensure that GENE is a factor with levels set by the summary.
# - Create a new column 'impact' that indicates whether protein_impact contains "ORF".
splicing_df <- splicing_df %>%
  mutate(
    absDelta = abs(deltapsi),
    GENE = factor(GENE, levels = levels(splicing_df_summary$GENE)),
    impact = ifelse(grepl("ORF", protein_impact, ignore.case = TRUE), "ORF", "Non-ORF")
  )

# Create the publication-ready plot:
p <- ggplot() +
  # Bar plot showing mean absolute deltapsi per gene, colored by the number of events
  geom_bar(
    data = splicing_df_summary, 
    aes(x = GENE, y = meanDelta, fill = factor(nEvents)),
    stat = "identity", 
    width = 0.7, 
    alpha = 0.8
  ) +
  # Overlay individual data points with jitter, mapping shape by the 'impact' column
  geom_jitter(
    data = splicing_df,
    aes(x = GENE, y = absDelta, shape = impact),
    width = 0.15,    # small horizontal jitter to reduce overplotting
    size = 2.5, 
    alpha = 0.6,     # adjust transparency
    color = "black"  # point border color
  ) +
  # Use a pleasing color palette for the bars (from RColorBrewer)
  scale_fill_brewer(palette = "Set2", name = "Number of Events") +
  # Manually set the shapes: for example, 16 (circle) for Non-ORF and 17 (triangle) for ORF
  scale_shape_manual(values = c("Non-ORF" = 16, "ORF" = 17), name = "Protein Impact") +
  # Publication-quality theme settings
  theme_bw(base_size = 14) +
  labs(
    x = "Gene ID",
    y = "|dPSI|",
    title = "Splicing related Genes in Spire"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

# Display the plot
print(p)


```


```{r splicing plots spire, warning=FALSE, message=FALSE, fig.width=16, fig.height=8}
spire_shared<-filter(spire_events$PSI, EVENT %in% splicing_df$EVENT)

# Add GENE names to the EVENT column
spire_shared <- spire_shared %>%
  mutate(EVENT = paste0(EVENT, " (", GENE, ")"))

spire_shared <- spire_shared %>%
  pivot_longer(
    cols = 7:10,  # Only selecting columns 7 to 10
    names_to = "Condition",
    values_to = "Value"
  ) %>%
  mutate(
    Group = case_when(
      Condition %in% names(spire_shared)[7:9] ~ "Control",
      Condition %in% names(spire_shared)[10:12] ~ "spireKO"
    )
  )


spire_shared$Group<-factor(spire_shared$Group, levels = c("Control","spireKO"))
# Plot with updated EVENT names
ggplot(spire_shared, aes(x = Group, y = Value, color = EVENT)) +
  geom_boxplot(size = 1,show.legend = FALSE) +
  labs(
    title = "Events of Splicing Proteins Spire",
    x = "",
    y = "PSI"
  ) +
  theme_minimal() +
  facet_wrap(~EVENT)
```



## FMNDKO

```{r splicing fmndko}

enrichdata<-enrichr(unique(c(differential_spire_exons$GENE, differential_spire_introns$GENE, differential_spire_alt$GENE,differential_fmndko_exons$GENE, differential_fmndko_introns$GENE, differential_fmndko_alt$GENE)), databases = c("GO_Biological_Process_2023","GO_Cellular_Component_2023","GO_Molecular_Function_2023"))

# Extract and format gene names
genes_splicing_1 <- enrichdata[[1]] %>%
  filter(str_detect(Term, "Splic")) %>%  # Select rows with 'splic' in Term
  pull(Genes) %>%                         # Extract Genes column
  str_split(";") %>%                      # Split gene names
  unlist() %>%                            # Flatten list
  str_to_lower() %>%                      # Convert to lowercase
  str_replace("^.", toupper) 
# Capitalize first letter
genes_splicing_2 <- enrichdata[[2]] %>%
  filter(str_detect(Term, "splic") | str_detect(Term, "Splic")) %>%  # Select rows with 'splic' in Term
  pull(Genes) %>%                         # Extract Genes column
  str_split(";") %>%                      # Split gene names
  unlist() %>%                            # Flatten list
  str_to_lower() %>%                      # Convert to lowercase
  str_replace("^.", toupper) 

genes_splicing<-unique(c(genes_splicing_1,genes_splicing_2))

combined_df <- bind_rows(
  exons_fmndko = differential_fmndko_exons,
  introns_fmndko = differential_fmndko_introns,
  .id = "source"  # This will be the new column indicating the origin
)
splicing_df<-filter(combined_df, GENE %in% genes_splicing)
splicing_df$protein_impact<-protein_impact$ONTO[match(splicing_df$EVENT, protein_impact$EventID)]



# Prepare the summary data frame, ensuring proper sorting and factor ordering
splicing_df_summary <- splicing_df %>%
  # Remove duplicate rows for the same gene and event (if applicable)
  distinct(GENE, EVENT, .keep_all = TRUE) %>%
  group_by(GENE) %>%
  summarise(
    nEvents   = n(),                                        # count unique events per gene
    meanDelta = mean(abs(deltapsi), na.rm = TRUE),          # mean of absolute deltapsi
    sdDelta   = sd(abs(deltapsi), na.rm = TRUE)             # standard deviation (for reference)
  ) %>%
  # First sort by the number of events (desc), then by mean absolute deltapsi (desc)
  arrange(desc(nEvents), desc(meanDelta)) %>%
  # Set the factor levels for GENE according to the sorted order
  mutate(GENE = factor(GENE, levels = unique(GENE)))

# Update the original data frame:
# - Create a column for the absolute value of deltapsi.
# - Ensure that GENE is a factor with levels set by the summary.
# - Create a new column 'impact' that indicates whether protein_impact contains "ORF".
splicing_df <- splicing_df %>%
  mutate(
    absDelta = abs(deltapsi),
    GENE = factor(GENE, levels = levels(splicing_df_summary$GENE)),
    impact = ifelse(grepl("ORF", protein_impact, ignore.case = TRUE), "ORF", "Non-ORF")
  )

# Create the publication-ready plot:
p <- ggplot() +
  # Bar plot showing mean absolute deltapsi per gene, colored by the number of events
  geom_bar(
    data = splicing_df_summary, 
    aes(x = GENE, y = meanDelta, fill = factor(nEvents)),
    stat = "identity", 
    width = 0.7, 
    alpha = 0.8
  ) +
  # Overlay individual data points with jitter, mapping shape by the 'impact' column
  geom_jitter(
    data = splicing_df,
    aes(x = GENE, y = absDelta, shape = impact),
    width = 0.15,    # small horizontal jitter to reduce overplotting
    size = 2.5, 
    alpha = 0.6,     # adjust transparency
    color = "black"  # point border color
  ) +
  # Use a pleasing color palette for the bars (from RColorBrewer)
  scale_fill_brewer(palette = "Set2", name = "Number of Events") +
  # Manually set the shapes: for example, 16 (circle) for Non-ORF and 17 (triangle) for ORF
  scale_shape_manual(values = c("Non-ORF" = 16, "ORF" = 17), name = "Protein Impact") +
  # Publication-quality theme settings
  theme_bw(base_size = 14) +
  labs(
    x = "Gene ID",
    y = "|dPSI|",
    title = "Splicing related Genes in FMNDKO"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

# Display the plot
print(p)


```

```{r splicing plots, warning=FALSE, message=FALSE, fig.width=16, fig.height=8}
fmndko_shared<-filter(fmndko_events$PSI, EVENT %in% splicing_df$EVENT)

# Add GENE names to the EVENT column
fmndko_shared <- fmndko_shared %>%
  mutate(EVENT = paste0(EVENT, " (", GENE, ")"))

fmndko_shared <- fmndko_shared %>%
  pivot_longer(
    cols = 7:10,  # Only selecting columns 7 to 10
    names_to = "Condition",
    values_to = "Value"
  ) %>%
  mutate(
    Group = case_when(
      Condition %in% names(fmndko_shared)[7:8] ~ "Control",
      Condition %in% names(fmndko_shared)[9:10] ~ "fmndko"
    )
  )


fmndko_shared$Group<-factor(fmndko_shared$Group, levels = c("Control","fmndko"))
# Plot with updated EVENT names
ggplot(fmndko_shared, aes(x = Group, y = Value, color = EVENT)) +
  geom_boxplot(size = 1,show.legend = FALSE) +
  labs(
    title = "Events of Splicing Proteins FMNDKO",
    x = "",
    y = "PSI"
  ) +
  theme_minimal() +
  facet_wrap(~EVENT)
```


# Nuclear Speckle Protein Splicing Correlation { .tabset}

Nuclear speckles genes were obtained from the Human Protein Atlas
Differentially spliced events were filtered to obtain those proteins localizing within speckles. I performed pearson correlation between those selected events. In adittion to the FDR threshold, they also need to have a |dPSI|>=0.1.

## Spire

```{r nuclear spckeles data hadnling spire}
set.seed(43)
speckles_proteins<-read_tsv("nuclear_speckles_protein_human_atlas.tsv")

spire_speckle_prots <- spire_events$PSI %>%
  filter((EVENT %in% differential_spire_exons$EVENT[
            tolower(differential_spire_exons$GENE) %in% tolower(speckles_proteins$Gene) & 
            abs(differential_spire_exons$deltapsi) >= 0.1]) |
         (EVENT %in% differential_spire_introns$EVENT[
            tolower(differential_spire_introns$GENE) %in% tolower(speckles_proteins$Gene) & 
            abs(differential_spire_introns$deltapsi) >= 0.1]) |
         (EVENT %in% differential_spire_alt$EVENT[
            tolower(differential_spire_alt$GENE) %in% tolower(speckles_proteins$Gene) & 
            abs(differential_spire_alt$deltapsi) >= 0.1]))




rownames(spire_speckle_prots)<-spire_speckle_prots$EVENT

spire_speckle_prots_t <- t(spire_speckle_prots[,7:12])

# Calculate correlation between genes
cor_matrix <- cor(spire_speckle_prots_t)

heatmap <- Heatmap(cor_matrix,
                   name = "Correlation",
                   clustering_distance_rows = "euclidean",  # Row clustering method
                   clustering_method_rows = "complete",     # Clustering method for rows
                   show_row_dend = TRUE,                    # Show row dendrogram
                   show_column_dend = TRUE,                 # Show column dendrogram
                   column_title = "Correlation Heatmap",    # Title for the heatmap
                   row_title = "Variables",                 # Title for rows
                   column_title_gp = gpar(fontsize = 14),   # Title formatting
                   row_title_gp = gpar(fontsize = 14)       # Row title formatting
)

# Draw the heatmap
draw(heatmap)

```

```{r table speckles spire}
# Render DataTable with enhancements

datatable(
  spire_speckle_prots,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```



## FMNDKO

```{r nuclear spckeles data hadnling fmndko}
set.seed(43)

fmndko_speckle_prots <- fmndko_events$PSI %>%
  filter((EVENT %in% differential_fmndko_exons$EVENT[
            tolower(differential_fmndko_exons$GENE) %in% tolower(speckles_proteins$Gene) & 
            abs(differential_fmndko_exons$deltapsi) >= 0.1]) |
         (EVENT %in% differential_fmndko_introns$EVENT[
            tolower(differential_fmndko_introns$GENE) %in% tolower(speckles_proteins$Gene) & 
            abs(differential_fmndko_introns$deltapsi) >= 0.1]) |
         (EVENT %in% differential_fmndko_alt$EVENT[
            tolower(differential_fmndko_alt$GENE) %in% tolower(speckles_proteins$Gene) & 
            abs(differential_fmndko_alt$deltapsi) >= 0.1]))




rownames(fmndko_speckle_prots)<-fmndko_speckle_prots$EVENT

fmndko_speckle_prots_t <- t(fmndko_speckle_prots[,7:10])

# Calculate correlation between genes
cor_matrix <- cor(fmndko_speckle_prots_t)

heatmap <- Heatmap(cor_matrix,
                   name = "Correlation",
                   clustering_distance_rows = "euclidean",  # Row clustering method
                   clustering_method_rows = "complete",     # Clustering method for rows
                   show_row_dend = TRUE,                    # Show row dendrogram
                   show_column_dend = TRUE,                 # Show column dendrogram
                   column_title = "Correlation Heatmap",    # Title for the heatmap
                   row_title = "Variables",                 # Title for rows
                   column_title_gp = gpar(fontsize = 14),   # Title formatting
                   row_title_gp = gpar(fontsize = 14)       # Row title formatting
)

# Draw the heatmap
draw(heatmap)

```

```{r table speckles fmndko}
# Render DataTable with enhancements


datatable(
  fmndko_speckle_prots,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

# Splicing Dynamics Heatmap Spire

From now, onwards, only events with |dPSI|>=0.1 are considered.

```{r heatmap pre-prep, results='hide', fig.show='hide'}

# Extract all events
spire_all <- filterEvents(
  spire_events,
  types = c("C1", "C2", "C3", "S", "MIC", "IR","Alt3","Alt5"),
  N = 3
)
```

```{r heatmap prep, results='hide', fig.show='hide'}
set.seed(44)

spire_unique <- spire_all$PSI %>% distinct(EVENT, .keep_all = TRUE) 

spire_unique<-spire_unique[-which(is.na(spire_unique$EVENT)),]
# Set rownames and convert to matrix
rownames(spire_unique) <- spire_unique$EVENT
spire_unique <- as.matrix(spire_unique[,7:12])

# Filter rows based on the final event lists and with abs(deltapsi)>=0.1
filtered_events <- unique(c(
  differential_spire_exons[abs(differential_spire_exons$deltapsi)>=0.1, "EVENT"],
  differential_spire_introns[abs(differential_spire_introns$deltapsi)>=0.1, "EVENT"]
  ))
spire_unique <- spire_unique[rownames(spire_unique) %in% filtered_events, ]

# Create a dendrogram for the columns
col_dend <- hclust(dist(t(spire_unique))) %>%
  as.dendrogram() %>%
  color_branches(k = 2) # Color branches with 3 clusters

# Check if row names contain "EX"
contains_EX <- grepl("EX", rownames(spire_unique))

# Define an improved color gradient for the heatmap
col_fun <- colorRamp2(
  c(0, 50, 100),
  c("purple", "white", "orange")
)

number_splices<-8


reference_table<-data.frame(EVENT=c(differential_fmndko_exons[abs(differential_fmndko_exons$deltapsi)>=0.1, "EVENT"],differential_fmndko_introns[abs(differential_fmndko_introns$deltapsi)>=0.1, "EVENT"],differential_spire_exons[abs(differential_spire_exons$deltapsi)>=0.1, "EVENT"],differential_spire_introns[abs(differential_spire_introns$deltapsi)>=0.1, "EVENT"]), GENE=c(differential_fmndko_exons[abs(differential_fmndko_exons$deltapsi)>=0.1, "GENE"],differential_fmndko_introns[abs(differential_fmndko_introns$deltapsi)>=0.1, "GENE"],differential_spire_exons[abs(differential_spire_exons$deltapsi)>=0.1, "GENE"],differential_spire_introns[abs(differential_spire_introns$deltapsi)>=0.1, "GENE"]))



ht<-Heatmap(
  spire_unique, name = "PSI",
  clustering_distance_rows = "pearson",
  col = col_fun,
  cluster_columns  = col_dend,
  column_title = "Oocytes  Samples at Different concentrations of ",
  row_title = "Splicing Events of |dPSI|>=0.1",
  column_names_rot = 45,
  column_dend_reorder = c(1:6),
  row_km = number_splices,
  row_dend_reorder = F,
  rect_gp = gpar(col = "white", lwd = 0.3),
  column_names_gp = gpar(fontsize = 16),
  show_row_names = FALSE,
  na_col = "transparent"

)
ht = draw(ht); clustered_events<-row_order(ht)
text_list<-list()
# Loop through clustered events and assign dynamic names
for (i in 1:length(clustered_events)) {
  enrich_events <- enrichr(
    reference_table$GENE[reference_table$EVENT %in% rownames(spire_unique)[clustered_events[[i]]]],
    databases = c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023")
  )

  # Assign the dynamic name and value
  text_list[[paste0("text", i)]] <- paste(
    enrich_events$GO_Cellular_Component_2023$Term[1],
    enrich_events$GO_Cellular_Component_2023$Term[2],
    enrich_events$GO_Cellular_Component_2023$Term[3],
    enrich_events$GO_Molecular_Function_2023$Term[1],
    enrich_events$GO_Molecular_Function_2023$Term[2],
    enrich_events$GO_Molecular_Function_2023$Term[3],
    sep = "; \n"
  )
}

protein_impact <- read.table("PROT_IMPACT-mm10-v2.3.tab", sep = "\t",
                             header = TRUE, stringsAsFactors = FALSE, fill = TRUE, quote = "")

events_for_impact <- rownames(spire_unique)

final_impact <- data.frame(EventID = rownames(spire_unique)) %>%  # Extract row names as EventID
  left_join(protein_impact %>% filter(EventID %in% events_for_impact), by = "EventID") %>%
  mutate(ONTO = ifelse(grepl("isoform", ONTO, ignore.case = TRUE), "-1", ONTO)) %>%
  mutate(ONTO = ifelse(grepl("UTR", ONTO, ignore.case = TRUE), "-2", ONTO)) %>%
  mutate(ONTO = case_when(
    grepl("ORF", ONTO, ignore.case = TRUE) & grepl("inclusion", ONTO, ignore.case = TRUE) ~ "2",
    grepl("ORF", ONTO, ignore.case = TRUE) & grepl("exclusion", ONTO, ignore.case = TRUE) ~ "1",
    TRUE ~ ONTO
  )) %>%
  mutate(ONTO = as.numeric(ONTO)) %>%
  replace_na(list(ONTO = 0))  # Fill missing ONTO values with 0


# Define the left annotation (points) with axis labels
left_annotation <- rowAnnotation(
  Impact = anno_points(
    final_impact$ONTO,
    width = unit(4, "cm"),  # Adjusted width
    size = unit(4, "mm"),   # Adjusted size of points
    axis_param = list(
      side = "top",
      at = -2:2, # Numeric values to label
      labels = c(
        "Regulatory (5´UTR)",
        "Alternative Isoform",
        "Unknown/NonCoding",
        "ORF disruption upon exclusion",
        "ORF disruption upon inclusion"
      ),
      labels_rot = 45, # Rotation of labels
      gp = gpar(fontsize = 15) # Text style
    )
  )
)

# Define the right annotation (foo)
right_annotation <- rowAnnotation(
  foo = anno_empty(
    border = FALSE,
    width = max_text_width(unlist(text_list)) + unit(4, "mm")
  )
)

```

### Heatmap showcasing the PSI of every candidate EX and IN

**Note**: Gene ontology of *Cellular Component* and *Biological Process* for each of the clustered blocks (splicing events that the *pearson* algorithm has found they behave similarly) appears on the right. The predicted impact on the protein appears on the left.

```{r heatmap_plot, fig.width=30, fig.height=30, echo=TRUE, out.width="100%"}
set.seed(44)

ht <- Heatmap(
  spire_unique,
  name = "PSI",
  clustering_distance_rows = "pearson",
  col = col_fun,
  cluster_columns = col_dend,
  column_title = "Oocytes Samples",
  column_title_gp = gpar(fontsize = 18, fontface = "bold"),
  row_title = "Splicing Events of |dPSI| >= 0.1",
  row_title_gp = gpar(fontsize = 28, fontface = "italic"),
  column_names_rot = 45,
  column_dend_reorder = c(1:6),
  row_km = number_splices,
  left_annotation = left_annotation,   # Add the points on the left
  right_annotation = right_annotation, # Add foo on the right
  row_dend_reorder = F,
  column_names_gp = gpar(fontsize = 18, fontface = "bold", col = "darkblue"),
  show_row_names = FALSE,
  row_gap = unit(6, "mm"), # Adjust the gap size between rows
  column_gap = unit(4, "mm"), # Adjust the gap size between columns
  heatmap_legend_param = list(
    title = "PSI Value",
    title_gp = gpar(fontsize = 30),
    labels_gp = gpar(fontsize = 30),
    legend_position = c(-20, 0),  # Move the legend closer to the center (x, y coordinates)
    legend_direction = "vertical" # Align the legend horizontally
  )
)

# Draw the heatmap
ht <- draw(ht)

# Decorate the "foo" annotation slices (on the right)
for (i in seq_along(clustered_events)) {
  decorate_annotation("foo", slice = i, {
    grid.rect(x = 0, width = unit(2, "mm"), gp = gpar(fill = i, col = NA), just = "left")
    grid.text(paste(text_list[[i]], collapse = "\n"), x = unit(4, "mm"), just = "left", gp = gpar(fontsize = 20))
  })
}

```

### Table of Events from Heatmap

**Note**: the *heatmap_cluster* column references to the heatmap row cluster (from top to bottom) that specific event belongs to.

```{r, table heatmap final}
row_order <- unlist(row_order(ht), use.names = F)
col_order <- unlist(column_order(ht), use.names = F)


cluster_index <- unlist(lapply(seq_along(row_order(ht)), function(i) rep(i, length(row_order(ht)[[i]]))))

# Reorder the original matrix based on clustering
clustered_matrix <- spire_unique[row_order, col_order]

final_impact_string<-protein_impact %>%
  filter(EventID %in% events_for_impact) %>%
  arrange(match(EventID, events_for_impact))
final_impact_string<- final_impact_string[match(rownames(clustered_matrix),final_impact_string$EventID),]


heatmap_dataframe<-tibble(EVENT=rownames(clustered_matrix), heatmap_cluster=cluster_index, protein_impact=final_impact_string$ONTO)
heatmap_dataframe$GENE<-reference_table$GENE[match(heatmap_dataframe$EVENT, reference_table$EVENT)]
heatmap_dataframe<-select(heatmap_dataframe, GENE, EVENT, protein_impact, heatmap_cluster)

heatmap_dataframe_spire<-heatmap_dataframe
# Render DataTable with enhancements
datatable(
  heatmap_dataframe,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))


```


### Same Heatmap but averaging PSI values from same condition

```{r heatmap_plot short, fig.width=30, fig.height=30, echo=TRUE, out.width="100%"}
set.seed(44)

spire_unique_short <- cbind(
  rowMeans(spire_unique[, 1:3]),
  rowMeans(spire_unique[, 4:6])
)

# Convert to matrix and print result
colnames(spire_unique_short) <- c("Control", "SpireKO")

spire_unique_short <- spire_unique_short[row_order,]

ht <- Heatmap(
  spire_unique_short,
  name = "PSI",
  col = col_fun,
  column_title = "Oocytes Samples",
  column_title_gp = gpar(fontsize = 24, fontface = "bold"),
  row_title = "Splicing Events of |dPSI| >= 0.1",
  row_title_gp = gpar(fontsize = 28, fontface = "italic"),
  column_names_rot = 45,
  column_dend_reorder = c(1:3),
  cluster_rows = FALSE,   # <-- This disables row clustering
  cluster_columns = FALSE, # Optionally disable column clustering if desired
  column_names_gp = gpar(fontsize = 18, fontface = "bold", col = "darkblue"),
  show_row_names = FALSE,
  row_gap = unit(6, "mm"),      # Adjust the gap size between rows
  left_annotation = left_annotation[row_order],   # Add the points on the left
  right_annotation = right_annotation[row_order],
  column_gap = unit(4, "mm"),   # Adjust the gap size between columns
  heatmap_legend_param = list(
    title = "PSI Value",
    title_gp = gpar(fontsize = 30),
    labels_gp = gpar(fontsize = 30),
    legend_position = c(-20, 0),  # Move the legend closer to the center (x, y coordinates)
    legend_direction = "vertical" # Align the legend vertically
  )
)

# Draw the heatmap
ht <- draw(ht)

# Draw the heatmap




```

### Enrichment of ORF Disrupted Genes { .tabset}

Note that interpretation should be done taking into account whether that specific ORF-disrupting events is being spliced in or out. This just may give an idea of the ontology of disrupted proteins.

```{r orf enrichr}
enrichdata<-enrichr(reference_table$GENE[reference_table$EVENT %in% final_impact$EventID[final_impact$ONTO %in% c(1,2)]], databases = c("GO_Biological_Process_2023","GO_Cellular_Component_2023","GO_Molecular_Function_2023"))



```

#### GO Biological Process

```{r orf bioprocess, fig.width=15, fig.height=8}
plotEnrich(enrichdata[[1]])

datatable(
  enrichdata[[1]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Cellular Component

```{r orf cellcomponent, fig.width=15, fig.height=8}
plotEnrich(enrichdata[[2]])
datatable(
  enrichdata[[2]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Molecular Function

```{r orf molfunction, fig.width=15, fig.height=8}
plotEnrich(enrichdata[[3]])
datatable(
  enrichdata[[3]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

# Splicing Dynamics Heatmap FMNDKO

```{r heatmap prep fmndko, results='hide', fig.show='hide'}
set.seed(44)

fmndko_all <- filterEvents(
  fmndko_events,
  types = c("C1", "C2", "C3", "S", "MIC", "IR","Alt3","Alt5"),
  N = 3
)

fmndko_unique <- fmndko_all$PSI %>% distinct(EVENT, .keep_all = TRUE) 

fmndko_unique<-fmndko_unique[-which(is.na(fmndko_unique$EVENT)),]
# Set rownames and convert to matrix
rownames(fmndko_unique) <- fmndko_unique$EVENT
fmndko_unique <- as.matrix(fmndko_unique[,7:10])

# Filter rows based on the final event lists and with abs(deltapsi)>=0.1
filtered_events <- unique(c(
  differential_fmndko_exons[abs(differential_fmndko_exons$deltapsi)>=0.1, "EVENT"],
  differential_fmndko_introns[abs(differential_fmndko_introns$deltapsi)>=0.1, "EVENT"]
  ))
fmndko_unique <- fmndko_unique[rownames(fmndko_unique) %in% filtered_events, ]

# Arrange columns based on metadata
# Create a dendrogram for the columns
col_dend <- hclust(dist(t(fmndko_unique))) %>%
  as.dendrogram() %>%
  color_branches(k = 2) # Color branches with 3 clusters

# Check if row names contain "EX"
contains_EX <- grepl("EX", rownames(fmndko_unique))

# Define an improved color gradient for the heatmap
col_fun <- colorRamp2(
  c(0, 50, 100),
  c("purple", "white", "orange")
)

number_splices<-6




ht<-Heatmap(
  fmndko_unique, name = "PSI",
  clustering_distance_rows = "pearson",
  col = col_fun,
  cluster_columns  = col_dend,
  column_title = "Oocytes  Samples at Different concentrations ",
  row_title = "Splicing Events of |dPSI|>=0.1",
  column_names_rot = 45,
  column_dend_reorder = c(1:4),
  row_km = number_splices,
  row_dend_reorder = F,
  rect_gp = gpar(col = "white", lwd = 0.3),
  column_names_gp = gpar(fontsize = 16),
  show_row_names = FALSE,
  na_col = "transparent"

)
ht = draw(ht); clustered_events<-row_order(ht)
text_list<-list()
# Loop through clustered events and assign dynamic names
for (i in 1:length(clustered_events)) {
  enrich_events <- enrichr(
    reference_table$GENE[reference_table$EVENT %in% rownames(fmndko_unique)[clustered_events[[i]]]],
    databases = c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023")
  )

  # Assign the dynamic name and value
  text_list[[paste0("text", i)]] <- paste(
    enrich_events$GO_Cellular_Component_2023$Term[1],
    enrich_events$GO_Cellular_Component_2023$Term[2],
    enrich_events$GO_Cellular_Component_2023$Term[3],
    enrich_events$GO_Molecular_Function_2023$Term[1],
    enrich_events$GO_Molecular_Function_2023$Term[2],
    enrich_events$GO_Molecular_Function_2023$Term[3],
    sep = "; \n"
  )
}

protein_impact <- read.table("PROT_IMPACT-mm10-v2.3.tab", sep = "\t",
                             header = TRUE, stringsAsFactors = FALSE, fill = TRUE, quote = "")

events_for_impact<-rownames(fmndko_unique)

final_impact <- data.frame(EventID = rownames(fmndko_unique)) %>%  # Extract row names as EventID
  left_join(protein_impact %>% filter(EventID %in% events_for_impact), by = "EventID") %>%
  mutate(ONTO = ifelse(grepl("isoform", ONTO, ignore.case = TRUE), "-1", ONTO)) %>%
  mutate(ONTO = ifelse(grepl("UTR", ONTO, ignore.case = TRUE), "-2", ONTO)) %>%
  mutate(ONTO = case_when(
    grepl("ORF", ONTO, ignore.case = TRUE) & grepl("inclusion", ONTO, ignore.case = TRUE) ~ "2",
    grepl("ORF", ONTO, ignore.case = TRUE) & grepl("exclusion", ONTO, ignore.case = TRUE) ~ "1",
    TRUE ~ ONTO
  )) %>%
  mutate(ONTO = as.numeric(ONTO)) %>%
  replace_na(list(ONTO = 0))  # Fill missing ONTO values with 0


# Define the left annotation (points) with axis labels
left_annotation <- rowAnnotation(
  Impact = anno_points(
    final_impact$ONTO,
    width = unit(4, "cm"),  # Adjusted width
    size = unit(4, "mm"),   # Adjusted size of points
    axis_param = list(
      side = "top",
      at = -2:2, # Numeric values to label
      labels = c(
        "Regulatory (5´UTR)",
        "Alternative Isoform",
        "Unknown/NonCoding",
        "ORF disruption upon exclusion",
        "ORF disruption upon inclusion"
      ),
      labels_rot = 45, # Rotation of labels
      gp = gpar(fontsize = 15) # Text style
    )
  )
)

# Define the right annotation (foo)
right_annotation <- rowAnnotation(
  foo = anno_empty(
    border = FALSE,
    width = max_text_width(unlist(text_list)) + unit(4, "mm")
  )
)

```

### Heatmap showcasing the PSI of every candidate EX and IN

**Note**: Gene ontology of *Cellular Component* and *Biological Process* for each of the clustered blocks (splicing events that the *pearson* algorithm has found they behave similarly) appears on the right. The predicted impact on the protein appears on the left.

```{r heatmap_plot fmndko, fig.width=30, fig.height=30, echo=TRUE, out.width="100%"}
set.seed(44)

ht <- Heatmap(
  fmndko_unique,
  name = "PSI",
  clustering_distance_rows = "pearson",
  col = col_fun,
  cluster_columns = col_dend,
  column_title = "Oocytes Samples",
  column_title_gp = gpar(fontsize = 24, fontface = "bold"),
  row_title = "Splicing Events of |dPSI| >= 0.1",
  row_title_gp = gpar(fontsize = 28, fontface = "italic"),
  column_names_rot = 45,
  column_labels = metadata_fmn2dko$condition,
  column_dend_reorder = c(1:4),
  row_km = number_splices,
  left_annotation = left_annotation,   # Add the points on the left
  right_annotation = right_annotation, # Add foo on the right
  row_dend_reorder = F,
  column_names_gp = gpar(fontsize = 18, fontface = "bold", col = "darkblue"),
  show_row_names = FALSE,
  row_gap = unit(6, "mm"), # Adjust the gap size between rows
  column_gap = unit(4, "mm"), # Adjust the gap size between columns
  heatmap_legend_param = list(
    title = "PSI Value",
    title_gp = gpar(fontsize = 30),
    labels_gp = gpar(fontsize = 30),
    legend_position = c(-20, 0),  # Move the legend closer to the center (x, y coordinates)
    legend_direction = "vertical" # Align the legend horizontally
  )
)

# Draw the heatmap
ht <- draw(ht)

# Decorate the "foo" annotation slices (on the right)
for (i in seq_along(clustered_events)) {
  decorate_annotation("foo", slice = i, {
    grid.rect(x = 0, width = unit(2, "mm"), gp = gpar(fill = i, col = NA), just = "left")
    grid.text(paste(text_list[[i]], collapse = "\n"), x = unit(4, "mm"), just = "left", gp = gpar(fontsize = 20))
  })
}

```

### Table of Events from Heatmap

**Note**: the *heatmap_cluster* column references to the heatmap row cluster (from top to bottom) that specific event belongs to.

```{r, table heatmap final fmndko}
row_order <- unlist(row_order(ht), use.names = F)
col_order <- unlist(column_order(ht), use.names = F)


cluster_index <- unlist(lapply(seq_along(row_order(ht)), function(i) rep(i, length(row_order(ht)[[i]]))))

# Reorder the original matrix based on clustering
clustered_matrix <- fmndko_unique[row_order, col_order]

final_impact_string<-protein_impact %>%
  filter(EventID %in% events_for_impact) %>%
  arrange(match(EventID, events_for_impact))
final_impact_string<- final_impact_string[match(rownames(clustered_matrix),final_impact_string$EventID),]


heatmap_dataframe<-tibble(EVENT=rownames(clustered_matrix), heatmap_cluster=cluster_index, protein_impact=final_impact_string$ONTO)
heatmap_dataframe$GENE<-reference_table$GENE[match(heatmap_dataframe$EVENT, reference_table$EVENT)]
heatmap_dataframe<-select(heatmap_dataframe, GENE, EVENT, protein_impact, heatmap_cluster)

# Render DataTable with enhancements
datatable(
  heatmap_dataframe,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))


```

### Same Heatmap but averaging PSI values from same condition

```{r heatmap_plot short fmndko, fig.width=30, fig.height=30, echo=TRUE, out.width="100%"}
set.seed(44)

fmndko_unique_short <- cbind(
  rowMeans(fmndko_unique[, 1:2]),
  rowMeans(fmndko_unique[, 3:4])
)

# Convert to matrix and print result
colnames(fmndko_unique_short) <- c("Control", "FMNDKO")

fmndko_unique_short <- fmndko_unique_short[row_order,]

ht <- Heatmap(
  fmndko_unique_short,
  name = "PSI",
  col = col_fun,
  column_title = "Oocytes Samples",
  column_title_gp = gpar(fontsize = 24, fontface = "bold"),
  row_title = "Splicing Events of |dPSI| >= 0.1",
  row_title_gp = gpar(fontsize = 28, fontface = "italic"),
  column_names_rot = 45,
  column_dend_reorder = c(1:3),
  cluster_rows = FALSE,   # <-- This disables row clustering
  cluster_columns = FALSE, # Optionally disable column clustering if desired
  column_names_gp = gpar(fontsize = 18, fontface = "bold", col = "darkblue"),
  show_row_names = FALSE,
  row_gap = unit(6, "mm"),      # Adjust the gap size between rows
  left_annotation = left_annotation[row_order],   # Add the points on the left
  right_annotation = right_annotation[row_order],
  column_gap = unit(4, "mm"),   # Adjust the gap size between columns
  heatmap_legend_param = list(
    title = "PSI Value",
    title_gp = gpar(fontsize = 30),
    labels_gp = gpar(fontsize = 30),
    legend_position = c(-20, 0),  # Move the legend closer to the center (x, y coordinates)
    legend_direction = "vertical" # Align the legend vertically
  )
)

# Draw the heatmap
ht <- draw(ht)

# Draw the heatmap




```

### Enrichment of ORF Disrupted Genes { .tabset}

Note that interpretation should be done taking into account whether that specific ORF-disrupting events is being spliced in or out. This just may give an idea of the ontology of disrupted proteins.

```{r orf enrichr fmndko}
enrichdata<-enrichr(unique(reference_table$GENE[reference_table$EVENT %in% final_impact$EventID[final_impact$ONTO %in% c(1,2)]]), databases = c("GO_Biological_Process_2023","GO_Cellular_Component_2023","GO_Molecular_Function_2023"))



```

#### GO Biological Process

```{r orf bioprocess fmndko, fig.width=15, fig.height=8}
plotEnrich(enrichdata[[1]])
datatable(
  enrichdata[[1]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Cellular Component

```{r orf cellcomponent fmndko, fig.width=15, fig.height=8}
plotEnrich(enrichdata[[2]])
datatable(
  enrichdata[[2]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```

#### GO Molecular Function

```{r orf molfunction fmndko, fig.width=15, fig.height=8}
plotEnrich(enrichdata[[3]])
datatable(
  enrichdata[[3]],
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
))
```



# System Settings

```{r session info}
sessionInfo()
```
